---
title: "Patient-Generated Health Data as a Predictor of Adverse Outcomes in Veterans"
subtitle: "Study 1 - Pre-Event PGHD Surveillance Cohort"
author: "Nick Cardamone"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: false
    theme: flatly
    highlight: tango
    code_folding: hide
    df_print: paged
    self_contained: true
    css: |
      body { font-family: Helvetica, Arial, sans-serif; font-size: 13px; }
      h1, h2, h3, h4 { font-family: Helvetica, Arial, sans-serif; }
      .table { font-size: 12px; }
      caption { font-size: 12px; font-style: italic; color: #555; }
      .fig-caption { font-size: 11px; color: #555; margin-top: 4px; margin-bottom: 16px; }
      .highlight-box { background: #f7f9fc; border-left: 4px solid #4575b4; padding: 10px 14px;
                       margin: 12px 0; font-size: 12px; }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5,
  dpi        = 300,
  dev        = "png",
  out.width  = "100%",
  cache      = FALSE
)

options(knitr.table.format = "html")

# ── Packages ──────────────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(tidyverse) # multitude of pacakges related to data cleaning
  library(arrow) # working with parquets
  library(data.table) # working with large dtasets
  library(pROC) # visualizing ROC curves
  library(broom) # convert statistical objects to tidy objects
  library(recipes) # making machine learning models
  library(gtsummary) # elegant tables
  library(gt) # elegant tables
  library(kableExtra) # html tables
  library(patchwork) # ordering visuals
  library(scales) # reformatting numbers
  library(sandwich) # cluster-robust standard errors
  library(lmtest) # linear model tests
  library(writexl) # saving to xlsx
})

for (pkg in c("xgboost", "glmnet", "PRROC", "downloadthis")) {
  if (!requireNamespace(pkg, quietly = TRUE))
    try(install.packages(pkg, repos = "https://cloud.r-project.org", quiet = TRUE), silent = TRUE)
}
if (requireNamespace("xgboost",      quietly = TRUE)) suppressPackageStartupMessages(library(xgboost)) # xgboost package
if (requireNamespace("glmnet",       quietly = TRUE)) suppressPackageStartupMessages(library(glmnet)) # logistic regression
if (requireNamespace("PRROC",        quietly = TRUE)) suppressPackageStartupMessages(library(PRROC)) # calcualte AUC
if (requireNamespace("downloadthis", quietly = TRUE)) suppressPackageStartupMessages(library(downloadthis)) # download button

# Set working directory 
base_dir <- "C:/Users/VHAPHICardaN/OneDrive - Department of Veterans Affairs/Desktop/Projects/OPS_Bressman-PGHDPred/pghdpred_deliverable"
setwd(base_dir)

# Save figures and output data
out_data    <- "data/pub"
out_figures <- "figures/pub"
for (d in c(out_data, out_figures)) if (!dir.exists(d)) dir.create(d, recursive = TRUE)

set.seed(2026L)
'%!in%' <- function(x, y) !('%in%'(x, y))
```

```{r theme, include=FALSE}
theme_science <- function(base_size = 9) {
  ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(
      text             = ggplot2::element_text(family = "Gill Sans MT"),
      plot.title       = ggplot2::element_text(size = base_size, face = "plain"),
      axis.title       = ggplot2::element_text(size = base_size),
      axis.text        = ggplot2::element_text(size = base_size - 1),
      strip.text       = ggplot2::element_text(size = base_size),
      legend.position  = "none",
      panel.grid.minor = ggplot2::element_blank(),
      panel.border     = ggplot2::element_rect(colour = "black", fill = NA, linewidth = 0.4),
      panel.background = ggplot2::element_rect(fill = "white", colour = NA)
    )
}

event_colors <- c(
  "ED Visit"  = "#d73027",
  "Inpatient" = "#4575b4",
  "Death"     = "#1a1a1a",
  "No Event"  = "#69b3a2"
)
feat_colors <- c(
  "Steps only"          = "#69b3a2",
  "Static only"         = "#4575b4",
  "Steps + static"      = "#d73027"
)
feat_labels <- c(
  model1_steps_only       = "Steps only",
  model2_static_only      = "Static only",
  model3_steps_and_static = "Steps + static"
)
```

```{r helpers, include=FALSE}
# Utility functions
# Take largest among two observations of the same measurement on the same day.
pmax_na <- function(...) {
  x <- pmax(..., na.rm = TRUE); x[is.infinite(x)] <- NA_real_; x
}
# Round to 5 digits
round_numeric <- function(df, digits = 5) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df %>% mutate(across(where(is.numeric), ~round(., digits)))
}
# Remove outliers among each measurement
remove_outliers <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE); q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  ifelse(x < (q1 - 1.5*iqr) | x > (q3 + 1.5*iqr), NA_real_, x)
}

# Format p-values
fmt_p <- function(p) ifelse(is.na(p), "--", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# Rolling helpers
roll_avg_with_days <- function(x, n) {
  x0 <- fifelse(is.na(x), 0, x)
  d  <- as.integer(!is.na(x))
  s  <- frollsum(x0, n = n, align = "right")
  k  <- frollsum(d,  n = n, align = "right")
  avg <- fifelse(k > 0L, s / k, NA_real_)
  list(avg = avg, days = k)
}

roll_sd <- function(x, n) {
  if (length(x) == 0) return(numeric(0))
  frollapply(x, n = n, FUN = function(w) {
    valid <- w[!is.na(w)]
    if (length(valid) >= 3L) sd(valid) else NA_real_
  }, align = "right")
}

# Outcome helper 
# Count days to event
calc_y_horizon <- function(days_to_event, horizon_days) {
  as.integer(!is.na(days_to_event) & days_to_event >= 1L & days_to_event <= horizon_days)
}

# Model builders 
# Dummy variables are pre-encoded as integers in modeling_data; recipes only
# handle imputation, normalization, and zero-variance removal.
# All categorical variables have their numeric variables set to "unknown". There were 38 cases where individuals had missinging medical condition information, those individuals were dropped. 
make_recipe_glmnet <- function(train_df) {
  recipe(Y ~ ., data = train_df) %>%
    step_impute_median(all_numeric_predictors()) %>% # impute median for numeric features
    step_normalize(all_numeric_predictors()) %>% # normalize numeric
    step_zv(all_predictors()) # drop any predictor that has no variance (i.e. all the same value)
}
make_recipe_xgboost <- function(train_df) {
  recipe(Y ~ ., data = train_df) %>%
    # imputation not needed because missingness is handled natively by xgboost
    step_normalize(all_numeric_predictors()) %>% # norm
    step_zv(all_predictors())
}

# Undersampling 
# We noticed we had considerably more non-event weeks than event weeks.
undersample_by_patient <- function(df, ratio = 3L, seed = 2026L) {
  set.seed(seed)
  df <- as_tibble(df)
  df %>%
    group_by(PatientICN) %>%
    group_modify(~{
      pos <- .x %>% filter(Y == 1); neg <- .x %>% filter(Y == 0)
      if (nrow(pos) == 0L) {
        if (nrow(neg) > 0L) neg[sample.int(nrow(neg), min(8L, nrow(neg))), ] else .x[0L, ]
      } else {
        n_neg <- min(nrow(pos) * ratio, nrow(neg))
        neg_s <- if (nrow(neg) > 0L && n_neg > 0L)
          neg[sample.int(nrow(neg), n_neg), ] else .x[0L, ]
        bind_rows(pos, neg_s)
      }
    }) %>% ungroup()
}

# Metrics
compute_global_metrics <- function(y, pred) {
  valid <- !is.na(y) & !is.na(pred)
  y <- y[valid]; pred <- pred[valid]
  roc_obj <- pROC::roc(response = y, predictor = pred, quiet = TRUE, direction = "<")
  auc_val <- as.numeric(pROC::auc(roc_obj))
  brier   <- mean((pred - y)^2) # brier score
  # area under precision recall curve
  pr_auc  <- tryCatch({
    PRROC::pr.curve(scores.class0 = pred[y == 1], scores.class1 = pred[y == 0],
                    curve = FALSE)$auc.integral
  }, error = function(e) NA_real_)
  eps   <- 1e-6
  logit_p <- log(pmax(eps, pred) / pmax(eps, 1 - pred))
  # slope
  cal_slope <- tryCatch(
    coef(suppressWarnings(glm(y ~ logit_p, family = binomial())))[["logit_p"]],
    error = function(e) NA_real_)
  brier_null <- mean(y) * (1 - mean(y))
  # brier skill score: better than brier given imbalanced data because its normalized against null models brier 
  bss        <- if (!is.na(brier_null) && brier_null > 0) 1 - brier / brier_null else NA_real_
  list(roc = roc_obj, auc = auc_val, pr_auc = pr_auc, brier = brier, bss = bss, cal_slope = cal_slope)
}

# 
make_calibration_table <- function(pred, y) {
  breaks <- unique(quantile(pred, probs = seq(0, 1, by = 0.1), na.rm = TRUE))
  tibble(predicted_prob = pred, actual = y) %>%
    mutate(prob_bin = cut(predicted_prob,
      breaks = breaks,
      include.lowest = TRUE)) %>%
    group_by(prob_bin) %>%
    summarise(n = n(), n_events = sum(actual),
              mean_predicted = mean(predicted_prob),
              observed_rate  = mean(actual), .groups = "drop") %>%
    round_numeric()
}

#  kableExtra re-render fix (VA/Windows: stale RStudio notebook cache dir)
# When the .Rproj.user/shared/notebooks chunk-cache directory is missing (e.g.
# after a new RStudio session), RStudio's print.shiny.tag.list fires a warning
# then an error trying to write HTML-dependency JSON to a stale path.
# Fix: capture the path from the warning, create the directory, retry once.
print.kableExtra <- function(x, ...) {
  dep_dir <- NULL
  withCallingHandlers(
    tryCatch(
      kableExtra:::print.kableExtra(x, ...),
      error = function(e) {
        if (grepl("cannot open the connection", conditionMessage(e), fixed = TRUE) &&
            !is.null(dep_dir)) {
          dir.create(dep_dir, recursive = TRUE, showWarnings = FALSE)
          kableExtra:::print.kableExtra(x, ...)
        } else {
          stop(e)
        }
      }
    ),
    warning = function(w) {
      if (grepl("cannot open file.*\\.Rproj\\.user", conditionMessage(w))) {
        m <- regmatches(conditionMessage(w),
                        regexpr("(?<=')[^']+(?=')", conditionMessage(w), perl = TRUE))
        if (length(m) && nchar(m) > 0L) dep_dir <<- dirname(m)
        invokeRestart("muffleWarning")
      }
    }
  )
}

# gtsummary UTF-8 fix (VA/Windows: invalid byte sequences crash gt HTML render) 
# Sanitizes all character columns in gtsummary internals before gt renders HTML
print.gtsummary <- function(x, ...) {
  safe_utf8 <- function(s) iconv(s, to = "UTF-8", sub = "")
  walk_df   <- function(obj) {
    if (is.data.frame(obj))
      dplyr::mutate(obj, dplyr::across(where(is.character), safe_utf8))
    else obj
  }
  x$table_body    <- walk_df(x$table_body)
  x$table_styling <- lapply(x$table_styling, walk_df)
  .orig <- get("print.gtsummary", envir = asNamespace("gtsummary"))
  .orig(x, ...)
}

# Feature importance labeling
classify_feature <- function(nm) {
  case_when(
    str_detect(nm, "k_steps_|baseline_k_steps|low_activity") ~ "PGHD / step features",
    str_detect(nm, "bmi_|map_|pulse|bun_|albumin_|leuk_|spo2")    ~ "Vitals & labs",
    str_detect(nm, "py2_inpat|py_ed|py_urgent|py_appt")           ~ "Healthcare utilization",
    str_detect(nm, "^py2_")                                        ~ "Comorbidity flags",
    str_detect(nm, "Age|gender_|race_|eth_|marital_|gisurh_|pg_G|pg_Unknown|season_|days_obs_remaining") ~ "Demographics",
    str_detect(nm, "cci_score")                                    ~ "Comorbidity flags",
    str_detect(nm, "mst_|smoker|tobacco")                          ~ "Social history",
    TRUE                                                            ~ "Other"
  )
}

# ── Feature display name lookup (used by missingness table, importance plots, coef tables) ──
feat_display_names <- c(
  # PGHD / step features
  baseline_k_steps_avg         = "Baseline avg. daily steps",
  baseline_k_steps_miss_days   = "Baseline avg. missing days/week",
  k_steps_7d_avg               = "7-day avg. daily steps",
  k_steps_7d_sd                = "7-day SD of daily steps",
  k_steps_7d_miss_days         = "Missing days/week (7-day)",
  k_steps_7d_all_missing       = "All days missing flag (7-day)",
  k_steps_7d_pct_from_baseline = "% change from baseline steps",
  k_steps_7d_avg_pctdiff       = "% change vs. prior 3-week avg.",
  low_activity_days_7d         = "Low-activity days/week (7-day)",
  low_activity_streak          = "Consecutive low-activity days",
  k_steps_14d_avg              = "14-day avg. daily steps",
  k_steps_14d_sd               = "14-day SD of daily steps",
  k_steps_21d_avg              = "21-day avg. daily steps",
  k_steps_21d_sd               = "21-day SD of daily steps",
  # Vitals & labs
  bmi_fitted_mean              = "BMI, fitted mean (kg/m\u00b2)",
  bmi_slope                    = "BMI trend slope",
  bmi_fitted_last              = "BMI, most recent fitted (kg/m\u00b2)",
  map_fitted_mean              = "MAP, fitted mean (mmHg)",
  map_fitted_last              = "MAP, most recent fitted (mmHg)",
  map_slope                    = "MAP trend slope",
  recent_pulse                 = "Heart rate, most recent (bpm)",
  bun_value_mean               = "BUN, mean (mg/dL)",
  bun_value_last               = "BUN, most recent (mg/dL)",
  albumin_value_mean           = "Albumin, mean (g/dL)",
  albumin_value_last           = "Albumin, most recent (g/dL)",
  leuk_value_mean              = "WBC, mean (10\u00b3/\u00b5L)",
  leuk_value_last              = "WBC, most recent (10\u00b3/\u00b5L)",
  # Demographics
  Age                          = "Age (years)",
  # Gender (ref = Male)
  gender_Female                = "Female (Male)",
  gender_Unknown               = "Unknown Sex (Male)",
  # Race (ref = White)
  race_BlackOrAfricanAmerican  = "Black or African American (White)",
  race_AmericanIndian          = "American Indian or Alaska Native (White)",
  race_Asian                   = "Asian (White)",
  race_PacificIslander         = "Native Hawaiian or Pacific Islander (White)",
  race_Multiracial             = "Multiracial (White)",
  race_Unknown                 = "Unknown Race (White)",
  # Ethnicity (ref = Not Hispanic or Latino)
  eth_HispanicOrLatino         = "Hispanic or Latino (Not Hispanic or Latino)",
  eth_Unknown                  = "Unknown Ethnicity (Not Hispanic or Latino)",
  # Marital status (ref = Married)
  marital_Divorced             = "Divorced (Married)",
  marital_NeverMarried         = "Never Married (Married)",
  marital_Widowed              = "Widowed (Married)",
  marital_Separated            = "Separated (Married)",
  marital_Unknown              = "Unknown Marital Status (Married)",
  # Rurality (ref = Urban)
  gisurh_Rural                 = "Rural (Urban)",
  gisurh_HighlyRural           = "Highly Rural (Urban)",
  gisurh_Insular               = "Insular (Urban)",
  gisurh_Unknown               = "Unknown Rurality (Urban)",
  # VA Priority Group - individual groups (ref = Group 1)
  pg_G2                        = "Priority Group 2 (Group 1)",
  pg_G3                        = "Priority Group 3 (Group 1)",
  pg_G4                        = "Priority Group 4 (Group 1)",
  pg_G5                        = "Priority Group 5 (Group 1)",
  pg_G6                        = "Priority Group 6 (Group 1)",
  pg_G7                        = "Priority Group 7 (Group 1)",
  pg_G8                        = "Priority Group 8 (Group 1)",
  pg_Unknown                   = "Unknown Priority Group (Group 1)",
  # Season (ref = Fall)
  season_Spring                = "Spring (Fall)",
  season_Summer                = "Summer (Fall)",
  season_Winter                = "Winter (Fall)",
  days_obs_remaining           = "Days remaining in follow-up",
  # Healthcare utilization
  py_ed                        = "ED visit, prior year",
  py_urgent                    = "Urgent care visit, prior year",
  py_appt_no_show              = "Appointment no-show, prior year",
  py2_inpat_any                = "Any inpatient admission, 2-year",
  py2_inpat_med_surg           = "Med./surgical admission, 2-year",
  py2_inpat_mental_health      = "Mental health admission, 2-year",
  py2_inpat_other              = "Other inpatient admission, 2-year",
  py2_inpat_nursing_home       = "Nursing home admission, 2-year",
  py2_IVC_n                    = "Community care stays, count (2-year)",
  py2_IVC_los                  = "Community care LOS, days (2-year)",
  py2_IVC                      = "Community care stay, any (2-year)",
  py2_mwi_score                = "Multimorbidity Weighted Index (MWI)",
  # Comorbidity flags
  py2_congestive_heart_failure = "Congestive heart failure, 2-year",
  py2_heart_arrhythmias        = "Cardiac arrhythmia, 2-year",
  py2_drug_alcohol_dependence  = "Substance use disorder, 2-year",
  py2_metastatic_cancer        = "Metastatic cancer, 2-year",
  py2_lung_severe_cancers      = "Lung/severe cancer, 2-year",
  py2_dialysis_status          = "Dialysis, 2-year",
  py2_housing_instability      = "Housing instability, 2-year",
  py2_financial_sdoh           = "Financial hardship, 2-year",
  py2_acute_renal_failure      = "Acute renal failure, 2-year",
  py2_artificial_openings      = "Artificial openings, 2-year",
  py2_chronic_pancreatitis     = "Chronic pancreatitis, 2-year",
  py2_chronic_ulcer_skin       = "Chronic skin/ulcer condition, 2-year",
  py2_drug_alcohol_psychosis   = "Drug/alcohol-induced psychosis, 2-year",
  py2_pregnancy                = "Pregnancy/peripartum condition, 2-year",
  py2_nonspecific_psychosocial = "Nonspecific psychosocial condition, 2-year",
  py2_dialysis_access          = "Dialysis access, 2-year",
  # Social history
  mst_Yes                      = "Military Sexual Trauma: Yes (No)",
  current_smoker               = "Current smoker",
  never_smoker                 = "Never smoker",
  cci_score                    = "Charlson Comorbidity Index (CCI)"
)
```

```{r data-load, include = FALSE, eval = FALSE}
# Load parquet saved from builder programs  

# Paths to find parquets
data_dir <- file.path(base_dir, "data", "input")

# Pre-flight check - fails if any upstream file is missing
path_cohort <- file.path(data_dir, "cohort_full_demo.parquet")
path_pghd   <- file.path(data_dir, "pghd_final_full.parquet") # very large
path_inpat  <- file.path(data_dir, "inpat_daily.parquet")
path_ed     <- file.path(data_dir, "ed_urgent_daily.parquet")
path_cond   <- file.path(data_dir, "patient_condition_flags_prior.parquet")
path_hcu    <- file.path(data_dir, "hc_util_prior.parquet")
path_vitals <- file.path(data_dir, "final_vitals_features_with_splines.parquet")
path_hf     <- file.path(data_dir, "final_health_factors_tobacco_mst.parquet")

missing_files <- Filter(
  Negate(file.exists),
  c(path_cohort, path_pghd, path_inpat, path_ed, path_cond, path_hcu, path_vitals, path_hf))
if (length(missing_files) > 0L)
  stop("Missing upstream parquet file(s):\n  ", paste(missing_files, collapse = "\n  "))

# Load parquets

cohort_full_demo  <- open_dataset(path_cohort) %>% collect()
.cohort_icns_vec  <- cohort_full_demo %>% distinct(PatientICN) %>% pull(PatientICN)

# only pull people for whom we have demogaphic data from the daily PGHD data stream
pghd_daily_raw    <- open_dataset(path_pghd) %>%
                       filter(PatientICN %in% .cohort_icns_vec) %>% 
                       select(PatientICN,
                              date,
                              date_first,
                              date_last,
                              obs_end,
                              ends_with("exercise_steps"), # keeping only steps and exercise m
                              ends_with("exercise_m"),
                              )  %>%
                       collect()

rm(.cohort_icns_vec); gc()
inpat_daily       <- open_dataset(path_inpat)  %>% collect()
ed_urgent_daily   <- open_dataset(path_ed)     %>% collect()
cond_flags        <- open_dataset(path_cond)   %>% collect()
hcu_prior         <- open_dataset(path_hcu)    %>% distinct() %>% collect()
vitals_features   <- open_dataset(path_vitals) %>% collect()
health_factors    <- open_dataset(path_hf)     %>%
                       select(-any_of("date_first")) %>% collect()

# Build static features table 

cohort_static <- cohort_full_demo %>%
  left_join(cond_flags,      by = "PatientICN") %>%
  left_join(hcu_prior,       by = "PatientICN") %>%
  left_join(vitals_features, by = "PatientICN") %>%
  left_join(health_factors,  by = "PatientICN") %>% 
  filter(!is.na(py2_acute_renal_failure)) # just drop people without condition data since there's so few.

as_tibble(cohort_static) %>%
  write_parquet("data/output/cohort_static.parquet")

# Convert date columns
inpat_daily     <- inpat_daily     %>% mutate(date = as.Date(date))
ed_urgent_daily <- ed_urgent_daily %>% mutate(date = as.Date(date))
```

```{r feature-engineering, include = FALSE, eval = FALSE}
# ══════════════════════════════════════════════════════════════════════════════
# STUDY 1 FEATURE ENGINEERING 
# ══════════════════════════════════════════════════════════════════════════════

# ── Step 0: Build cohort_daily ────────────────────────────────────────────────
# pghd_daily_raw is pre-filtered to cohort patients in data-load (Arrow push-down),
# so no semi_join needed here - filter directly to the observation window.
cohort_daily <- pghd_daily_raw %>%
  mutate(date       = as.Date(date),
         date_first = as.Date(date_first),
         obs_end    = as.Date(obs_end)) %>%
  filter(!is.na(date_first),
         date_first >= as.Date("2022-01-01"),       # inception lower bound
         date_first <= as.Date("2025-03-16"),        # enrollment cutoff
         date >= date_first, date <= obs_end) %>%   # trim to obs window
  inner_join(cohort_static %>% select(PatientICN), by = "PatientICN") %>%
  left_join(cohort_full_demo %>% select(PatientICN, dod),             by = "PatientICN") %>%
  left_join(inpat_daily   %>% select(PatientICN, date, inpat_any),   by = c("PatientICN","date")) %>%
  left_join(ed_urgent_daily %>% select(PatientICN, date, ed),         by = c("PatientICN","date")) %>%
  mutate(inpat_any = coalesce(inpat_any, 0L),
         ed        = coalesce(ed, 0L))

gc()
```
```{r, include = FALSE, eval = FALSE}
# Outlier removal (function created above)
cohort_daily <- cohort_daily %>%
  mutate(across(
    # this will be relevant if we ever include more wearable features other than steps and m
    c(starts_with("ah_das_"), 
      starts_with("fit_das_"), 
      starts_with("gar_das_"),
      starts_with("ah_wo_"),  
      starts_with("fit_wo_"),  
      starts_with("gar_wo_"),
      starts_with("ah_sleep_"),
      starts_with("fit_sleep_"),
      starts_with("gar_sleep_")),
    remove_outliers))

# Device coalescing 
cohort_steps <- cohort_daily %>%
  transmute(
    PatientICN, date = as.Date(date),
    date_first = as.Date(date_first), 
    obs_end = as.Date(obs_end), 
    dod = as.Date(dod),
    ed, inpat_any,
    # take highest non-NA value from Fitbit and Garmin
    fit_garmin_steps     = pmax_na(fit_das_exercise_steps, gar_das_exercise_steps),
    apple_steps          = ah_das_exercise_steps,
    fit_garmin_exercise_m = pmax_na(fit_das_exercise_m, gar_das_exercise_m),
    apple_exercise_m     = ah_das_exercise_m
  )
```
```{r, include = FALSE, eval = FALSE}
# Build dynamic_daily 
dynamic_daily <- cohort_steps %>%
  group_by(PatientICN) %>% arrange(date) %>%
  mutate(
    daily_steps     = coalesce(fit_garmin_steps,      apple_steps),
    daily_exercise_m = coalesce(fit_garmin_exercise_m, apple_exercise_m)
  ) %>% ungroup() %>%
  mutate(across(c(daily_steps, daily_exercise_m), ~if_else(. == 0, NA_real_, .)),
         died = if_else(!is.na(dod) & as.Date(dod) == date, 1L, 0L),
         days_since_first = as.integer(difftime(date, as.Date(date_first), units = "days")))

# Record first event per person - if inpatient admission
event_types <- dynamic_daily %>%
  group_by(PatientICN) %>% arrange(date) %>%
  mutate(had_event = as.integer(ed == 1 | inpat_any == 1 | died == 1)) %>%
  summarise(
    first_event_date = { d <- date[had_event==1]; if (length(d)==0) as.Date(NA) else min(d) },
    first_event_type = {
      d <- date[had_event==1]
      if (length(d)==0) NA_character_
      else {
        fe <- min(d); r <- pick(everything()) %>% filter(date == fe)
        case_when(r$died[1]==1 ~ "Death", r$inpat_any[1]==1 ~ "Inpatient", TRUE ~ "ED Visit")
      }
    }, .groups = "drop")
```

```{r daily-processing, include = FALSE, eval = FALSE}
# Join event information to daily step/ exercise m data
dynamic_daily <- dynamic_daily %>%
  left_join(event_types, by = "PatientICN")

# Step imputation (OLS: steps ~ exercise_m)
dt_daily <- as.data.table(dynamic_daily)
dt_daily[, date := as.Date(date)]; dt_daily[, obs_end := as.Date(obs_end)]
setorder(dt_daily, PatientICN, date)

imp_train <- dt_daily[!is.na(daily_steps) & !is.na(daily_exercise_m),
                      .(steps = as.numeric(daily_steps), exercise_m = as.numeric(daily_exercise_m))]
step_imp_model <- if (nrow(imp_train) >= 50L)
  suppressWarnings(lm(steps ~ exercise_m, data = as.data.frame(imp_train))) else NULL

dt_daily[, daily_steps_imp := as.numeric(daily_steps)]
ni <- which(is.na(dt_daily$daily_steps_imp) & !is.na(dt_daily$daily_exercise_m))
if (length(ni) > 0L && !is.null(step_imp_model)) {
  pred_imp <- predict(step_imp_model,
    newdata = data.frame(exercise_m = as.numeric(dt_daily$daily_exercise_m[ni])))
  dt_daily$daily_steps_imp[ni] <- pmax(0, pred_imp)
}
```

```{r rolling features, include = FALSE, eval = FALSE}
# Rolling features
dt_daily[, c("steps_7d_avg","steps_7d_days") := {
  r <- roll_avg_with_days(daily_steps_imp, 7L); list(r$avg, r$days)
}, by = PatientICN]
dt_daily[, k_steps_7d_avg       := steps_7d_avg / 1000]
dt_daily[, k_steps_7d_miss_days := as.integer(7L - fifelse(is.na(steps_7d_days), 0L, steps_7d_days))]
dt_daily[, k_steps_7d_all_missing := fifelse(is.na(steps_7d_days), 0L, as.integer(steps_7d_days == 0L))]
dt_daily[, k_steps_7d_sd := roll_sd(daily_steps_imp, 7L) / 1000, by = PatientICN]

# Low activity streak 
dt_daily[, is_low_activity := as.integer(!is.na(daily_steps_imp) & daily_steps_imp < 1000)]
dt_daily[, low_activity_streak := {
  s <- integer(length(is_low_activity)); cur <- 0L
  for (i in seq_along(is_low_activity)) {
    if (is.na(is_low_activity[i])) { s[i] <- NA_integer_; cur <- 0L
    } else if (is_low_activity[i] == 1L) { cur <- cur + 1L; s[i] <- cur
    } else { s[i] <- 0L; cur <- 0L }
  }; s
}, by = PatientICN]
dt_daily[, low_activity_days_7d := frollsum(is_low_activity, n = 7L, align = "right"), by = PatientICN]

dt_daily[, c("steps_14d_avg","steps_14d_days") := {
  r <- roll_avg_with_days(daily_steps_imp, 14L); list(r$avg, r$days)
}, by = PatientICN]
dt_daily[, k_steps_14d_avg := steps_14d_avg / 1000]
dt_daily[, k_steps_14d_sd  := roll_sd(daily_steps_imp, 14L) / 1000, by = PatientICN]

dt_daily[, c("steps_21d_avg","steps_21d_days") := {
  r <- roll_avg_with_days(daily_steps_imp, 21L); list(r$avg, r$days)
}, by = PatientICN]
dt_daily[, k_steps_21d_avg := steps_21d_avg / 1000]
dt_daily[, k_steps_21d_sd  := roll_sd(daily_steps_imp, 21L) / 1000, by = PatientICN]

# Patient-relative calendar weeks (anchored to each patient's first upload)
dt_daily[, first_upload_date := min(date),                                         by = PatientICN]
dt_daily[, patient_week_index      := (as.integer(date - first_upload_date) %/% 7L) + 1L]
dt_daily[, weekly_end_date         := first_upload_date + patient_week_index * 7L - 1L]
```

```{r weekly-aggregation, include = FALSE, eval = FALSE}
weekly_cols <- c("date","obs_end",
  "k_steps_7d_avg","k_steps_7d_miss_days","k_steps_7d_all_missing","k_steps_7d_sd",
  "low_activity_days_7d","low_activity_streak",
  "k_steps_14d_avg","k_steps_14d_sd","k_steps_21d_avg","k_steps_21d_sd",
  "first_upload_date","weekly_end_date")

# One row per patient per patient-relative calendar week (last qualifying day of each week)
dt_weekly <- dt_daily[
  !is.na(k_steps_7d_avg),
  .SD[which.max(date)],
  by = .(PatientICN, patient_week_index),
  .SDcols = weekly_cols]
setnames(dt_weekly, "patient_week_index", "weekly_interval_index")
setorder(dt_weekly, PatientICN, weekly_interval_index)

# Percent-change features (vs previous 3 qualifying windows; winsorized to plus or minus 500%)
dt_weekly[, `:=`(
  lag1 = shift(k_steps_7d_avg, 1L),
  lag2 = shift(k_steps_7d_avg, 2L),
  lag3 = shift(k_steps_7d_avg, 3L)
), by = PatientICN]
dt_weekly[, prev3 := rowMeans(cbind(lag1, lag2, lag3), na.rm = TRUE)]
dt_weekly[is.nan(prev3), prev3 := NA_real_]
dt_weekly[, k_steps_7d_avg_pctdiff := {
  raw <- 100 * (k_steps_7d_avg - prev3) / prev3
  raw <- fifelse(!is.na(prev3) & prev3 >= 0.10, raw, NA_real_)
  pmax(pmin(raw, 500), -500)
}]
dt_weekly[, c("lag1","lag2","lag3","prev3") := NULL]

```

```{r consort-pipeline, include = FALSE, eval = FALSE}
# Load cohort parameters from single source 
source("cohort_config.R")

inception_cutoff <- as.Date("2022-01-01")

max_follow_up_days <- 52L * 7L        # 364 days = 52 weeks
baseline_weeks <- 6L
min_total_weeks <- baseline_weeks + 1L   # baseline + ≥1 observation week
.today            <- Sys.Date()
data_end_raw      <- .today                          # overwritten in pipeline
last_observable   <- data_end_raw  - max_horizon
followup_cutoff   <- last_observable - max_follow_up_days

# Human-readable labels (overwritten in pipeline once real data_end_raw is known)
data_start_date        <- format(inception_cutoff,  "%B %d, %Y")
data_end_date          <- format(data_end_raw,      "%B %d, %Y")
followup_cutoff_label  <- format(followup_cutoff,   "%B %d, %Y")

# Merge event types 
dt_analysis <- merge(dt_weekly, event_types, by = "PatientICN", all.x = TRUE)
dt_analysis[, days_to_event := as.integer(first_event_date - weekly_end_date)]
dt_analysis[is.na(first_event_type), first_event_type := "No Event"]
```

```{r, include = FALSE, eval = FALSE}
# remove Veteran with no ICN recorded
dt_analysis <- dt_analysis %>% 
  filter(PatientICN != "no ICN")
```

```{r, include = FALSE, eval = FALSE}
# CONSORT PIPELINE - capture N patients and N person-weeks at each step

# Recompute derived dates now that data is loaded
data_end_raw          <- max(as.Date(pghd_daily_raw$date), na.rm = TRUE)
last_observable       <- data_end_raw  - max_horizon # Subtract 30 days
followup_cutoff       <- last_observable - max_follow_up_days
followup_cutoff_label <- format(followup_cutoff, "%B %d, %Y") 
data_start_date       <- format(inception_cutoff, "%B %d, %Y")
data_end_date         <- format(data_end_raw,     "%B %d, %Y")

# Save inline text parameters for fast reload 
tibble(
  data_start_date       = data_start_date,
  data_end_date         = data_end_date,
  followup_cutoff_label = followup_cutoff_label,
  n_pdays_with_steps    = as.integer(sum(!is.na(dt_daily$fit_garmin_steps) | !is.na(dt_daily$apple_steps))),
  pct_fitgarmin         = round(100 * mean(!is.na(dt_daily$fit_garmin_steps), na.rm = TRUE), 1)
) %>% write_parquet("data/output/inline_params.parquet")
```

```{r, include = FALSE, eval = FALSE}
# Step 1: enrolled PGHD cohort - first upload 2022-01-01 through 2025-03-16
# (inception lower bound; upper bound ensures ≥30d potential observation)
dt_analysis <- dt_analysis[
  first_upload_date >= as.Date("2022-01-01") & first_upload_date <= as.Date("2025-03-16")]
n1_pts <- uniqueN(dt_analysis$PatientICN)
n1_wks <- nrow(dt_analysis)

# Step 2: require ≥1 qualifying week in baseline (weeks 1–6) and ≥1 qualifying observation week (week 7+)
n3_before <- uniqueN(dt_analysis$PatientICN)
has_base  <- dt_analysis[weekly_interval_index <= 6L, .(has = TRUE), by = PatientICN]
has_obs   <- dt_analysis[weekly_interval_index >= 7L, .(has = TRUE), by = PatientICN]
dt_analysis <- dt_analysis[
  PatientICN %in% has_base$PatientICN & PatientICN %in% has_obs$PatientICN]
n3_pts <- uniqueN(dt_analysis$PatientICN)
n3_wks <- nrow(dt_analysis)

# Compute 6-week baseline metrics
baseline_metrics <- dt_analysis[weekly_interval_index <= 6L,
  .(baseline_k_steps_avg = mean(k_steps_7d_avg, na.rm = TRUE),
    baseline_k_steps_miss_days = mean(k_steps_7d_miss_days, na.rm = TRUE),
    baseline_end_date = max(weekly_end_date)),
  by = PatientICN]
dt_analysis <- merge(dt_analysis, baseline_metrics, by = "PatientICN", all.x = TRUE)

# Step 3b: require event-free baseline - exclude patients whose composite event
# occurred during weeks 1–6.  These patients have no event-free baseline to
# anchor predictions on; without this step they would be silently dropped inside
# Step 4 (censor_date < all observation-week end dates → 0 rows retained).
event_in_baseline <- dt_analysis[
  !is.na(first_event_date) & first_event_date <= first_upload_date + 41L,
  unique(PatientICN)]
dt_analysis <- dt_analysis[PatientICN %!in% event_in_baseline]
n3b_pts <- uniqueN(dt_analysis$PatientICN)
n3b_wks <- nrow(dt_analysis)

# Move to observation period (weeks 7+) and compute pct-from-baseline
dt_analysis <- dt_analysis[!is.na(baseline_k_steps_avg) & weekly_interval_index >= 7L]
dt_analysis[, k_steps_7d_pct_from_baseline :=
              100 * (k_steps_7d_avg - baseline_k_steps_avg) / baseline_k_steps_avg]

# Step 4: event classification + censoring at event or max_follow_up_days (intervals removed, patients same)
# event_within_window is extended by max_horizon so events in the 30-day ascertainment
# headroom reserved by followup_cutoff = last_observable - max_follow_up_days are captured.
# This ensures labels at the final qualifying observation week are fully confirmed.
dt_analysis[, obs_window_end  := first_upload_date + max_follow_up_days]
dt_analysis[, event_within_window := !is.na(first_event_date) & first_event_date <= obs_window_end + max_horizon]
dt_analysis[, censor_date := fifelse(
  event_within_window,
  first_event_date,
  obs_window_end)]

# Save pre-step5 snapshot BEFORE censoring filter (for diagnostics) 
as_tibble(dt_analysis) %>%
  write_parquet("data/output/dt_analysis_pre_step4.parquet")

n4_before <- uniqueN(dt_analysis$PatientICN)
dt_analysis <- dt_analysis[weekly_end_date <= censor_date]
dt_analysis[event_within_window == FALSE,
  `:=`(first_event_date = NA_Date_, days_to_event = NA_integer_, first_event_type = "No Event")]
n4_pts <- uniqueN(dt_analysis$PatientICN)    # n4_before - n4_pts had their event before their first obs-week end
n4_wks <- nrow(dt_analysis)                  # fewer intervals

# Save CONSORT counts for fast reload 
tibble(
  n1_pts = n1_pts, n1_wks = n1_wks,
  n3_pts = n3_pts, n3_wks = n3_wks,
  n3b_pts = n3b_pts, n3b_wks = n3b_wks,
  n4_before = n4_before, n4_pts = n4_pts, n4_wks = n4_wks
) %>% write_parquet("data/output/consort_counts.parquet")
```
```{r, include = FALSE, eval = FALSE}

# ── Patient-level summary (for Table 1 and Figure 2) ──────────────────────────
patient_summary <- dt_analysis[,
  .(first_event_type       = first(first_event_type),
    first_upload_date       = first(first_upload_date),
    first_event_date       = first(first_event_date),
    days_to_event_from_start = as.integer(difftime(first(first_event_date),
                                                   first(first_upload_date), units = "days")),
    total_weeks            = .N,
    baseline_k_steps_avg   = first(baseline_k_steps_avg),
    baseline_k_steps_miss  = first(baseline_k_steps_miss_days),
    obs_k_steps_avg        = mean(k_steps_7d_avg,         na.rm = TRUE),
    obs_k_steps_sd         = sd(k_steps_7d_avg,           na.rm = TRUE),
    obs_miss_days_avg      = mean(k_steps_7d_miss_days,   na.rm = TRUE),
    obs_pct_baseline_avg   = mean(k_steps_7d_pct_from_baseline, na.rm = TRUE)),
  by = PatientICN]

# Join static features into patient summary for Table 1
tbl1_data <- as_tibble(patient_summary) %>%
  left_join(cohort_static %>% select(
    PatientICN, Age, Gender, Race, Ethnicity,
    MaritalStatus, GISURH, PriorityGroup, cci_score,
    starts_with("py2_"), py_ed, py_urgent, py_appt_no_show,
    bmi_fitted_mean, bmi_slope, map_fitted_mean, recent_pulse,
    bun_value_mean, albumin_value_mean, leuk_value_mean,
    mst_status, tobacco_status), by = "PatientICN") %>%
  mutate(
    first_event_type = factor(first_event_type,
      levels = c("Death","Inpatient","ED Visit","No Event")),
    had_event = as.integer(first_event_type != "No Event"),
    Gender        = factor(Gender),
    Race          = factor(Race),
    MaritalStatus = relevel(factor(MaritalStatus), ref = "MARRIED"), # set reference groups
    GISURH        = relevel(factor(GISURH), ref = "U"),
    mst_status    = factor(mst_status),
    tobacco_status = factor(tobacco_status),
    py2_IVC_n   = if_else(is.numeric(py2_IVC_n)   & py2_IVC_n   == 0, NA_real_, as.numeric(py2_IVC_n)),
    py2_IVC_los = if_else(is.numeric(py2_IVC_los) & py2_IVC_los == 0, NA_real_, as.numeric(py2_IVC_los)),
    py_ed         = factor(as.integer(py_ed)),
    py_urgent     = factor(as.integer(py_urgent)),
    py_appt_no_show = factor(as.integer(py_appt_no_show))
  )

dt_analysis %>% write_parquet("data//output//dt_analysis.parquet")
patient_summary %>% write_parquet("data//output//patient_summary.parquet")
tbl1_data %>% write_parquet("data//output//tbl1_data.parquet")

# ── Save snapshots for CONSORT diagnostics ────────────────────────────────────
# Save dt_weekly (post-rolling, pre-CONSORT) for re-running diagnostic filters
as_tibble(dt_weekly) %>% write_parquet("data/output/dt_weekly_snapshot.parquet")
as_tibble(event_types) %>% write_parquet("data/output/event_types_snapshot.parquet")

```

```{r train-test, eval = FALSE}
dt_analysis <- open_dataset("data//output//dt_analysis.parquet") %>% collect()
cohort_static <- open_dataset("data//output//cohort_static.parquet") %>% collect()


# Build modeling_data
modeling_data <- as_tibble(dt_analysis) %>%
  inner_join(cohort_static %>% select(
    PatientICN, Age, Gender, Race, Ethnicity, MaritalStatus, 
    GISURH, PriorityGroup,
    starts_with("py2_"), cci_score, 
    py_ed, py_urgent, py_appt_no_show,
    bmi_slope, bmi_fitted_mean, bmi_fitted_last,
    map_slope, map_fitted_mean, map_fitted_last, recent_pulse,
    bun_value_mean, bun_value_last, albumin_value_mean, albumin_value_last,
    leuk_value_mean, leuk_value_last, mst_status, tobacco_status),
    by = "PatientICN") %>%
  mutate(
    Y  = calc_y_horizon(days_to_event, 30L),  # placeholder; overridden in fit function
    # Manual dummy variables - reference groups explicit in column names ────
    # Gender (ref = Male; OMOP gender concept names are uppercase)
    gender_Female                = as.integer(!is.na(Gender) & toupper(Gender) == "FEMALE"),
    gender_Unknown               = as.integer(is.na(Gender) | !toupper(Gender) %in% c("MALE","FEMALE")),
    # Race (ref = White; OMOP concept names mixed-case)
    race_BlackOrAfricanAmerican  = as.integer(!is.na(Race) & Race == "Black or African American"),
    race_AmericanIndian          = as.integer(!is.na(Race) & Race == "American Indian or Alaska Native"),
    race_Asian                   = as.integer(!is.na(Race) & Race == "Asian"),
    race_PacificIslander         = as.integer(!is.na(Race) & Race == "Native Hawaiian or Other Pacific Islander"),
    race_Multiracial             = as.integer(!is.na(Race) & str_detect(tolower(as.character(Race)), "multiple|multi")),
    race_Unknown                 = as.integer(
      is.na(Race) |
      (!Race %in% c("White","Black or African American","American Indian or Alaska Native",
                    "Asian","Native Hawaiian or Other Pacific Islander") &
       !str_detect(tolower(coalesce(as.character(Race), "")), "multiple|multi"))),
    # Ethnicity (ref = Not Hispanic or Latino)
    eth_HispanicOrLatino         = as.integer(!is.na(Ethnicity) & Ethnicity == "Hispanic or Latino"),
    eth_Unknown                  = as.integer(
      is.na(Ethnicity) | !Ethnicity %in% c("Hispanic or Latino","Not Hispanic or Latino")),
    # Marital status (ref = MARRIED; raw values are uppercase from VA SPatient)
    marital_Divorced             = as.integer(!is.na(MaritalStatus) & toupper(MaritalStatus) == "DIVORCED"),
    marital_NeverMarried         = as.integer(!is.na(MaritalStatus) & toupper(MaritalStatus) %in% c("NEVER MARRIED","SINGLE")),
    marital_Widowed              = as.integer(!is.na(MaritalStatus) & toupper(MaritalStatus) == "WIDOWED"),
    marital_Separated            = as.integer(!is.na(MaritalStatus) & toupper(MaritalStatus) == "SEPARATED"),
    marital_Unknown              = as.integer(
      is.na(MaritalStatus) |
      !toupper(MaritalStatus) %in% c("MARRIED","DIVORCED","NEVER MARRIED","SINGLE","WIDOWED","SEPARATED")),
    # Rurality (ref = U = Urban; raw values are single letter codes)
    gisurh_Rural                 = as.integer(!is.na(GISURH) & toupper(GISURH) == "R"),
    gisurh_HighlyRural           = as.integer(!is.na(GISURH) & toupper(GISURH) == "H"),
    gisurh_Insular               = as.integer(!is.na(GISURH) & toupper(GISURH) == "I"),
    gisurh_Unknown               = as.integer(is.na(GISURH) | !toupper(GISURH) %in% c("U","R","H","I")),
    # VA Priority Group individual groups (ref = Group 1; raw = "Group N: ...")
    .pg_num                      = suppressWarnings(as.integer(str_extract(as.character(PriorityGroup), "[0-9]+"))),
    pg_G2                        = as.integer(!is.na(.pg_num) & .pg_num == 2L),
    pg_G3                        = as.integer(!is.na(.pg_num) & .pg_num == 3L),
    pg_G4                        = as.integer(!is.na(.pg_num) & .pg_num == 4L),
    pg_G5                        = as.integer(!is.na(.pg_num) & .pg_num == 5L),
    pg_G6                        = as.integer(!is.na(.pg_num) & .pg_num == 6L),
    pg_G7                        = as.integer(!is.na(.pg_num) & .pg_num == 7L),
    pg_G8                        = as.integer(!is.na(.pg_num) & .pg_num == 8L),
    pg_Unknown                   = as.integer(is.na(.pg_num) | !.pg_num %in% 1:8),
    # Season (ref = Fall; using lubridate::month)
    .mo                          = lubridate::month(weekly_end_date),
    season_Spring                = as.integer(.mo %in% 3:5),
    season_Summer                = as.integer(.mo %in% 6:8),
    season_Winter                = as.integer(.mo %in% c(12L, 1L, 2L)),
    # MST (ref = No/not reported)
    mst_Yes                      = as.integer(toupper(as.character(mst_status)) == "YES"),
    # Tobacco (binary flags; mutually non-exclusive - no reference encoding needed)
    current_smoker               = as.integer(tolower(as.character(tobacco_status)) == "current smoker"),
    never_smoker                 = as.integer(tolower(as.character(tobacco_status)) == "never smoker"),
    days_obs_remaining           = as.integer((obs_window_end + max_horizon) - weekly_end_date)
  )

# ── Drop columns ──────────────────────────────────────────────────────────────
drop_cols <- c("PatientICN","date","weekly_end_date","obs_end","first_event_date",
               "first_upload_date",
               "first_event_type","days_to_event","event_within_window","censor_date",
               "obs_window_end","weekly_interval_index","days_obs_remaining","baseline_end_date","tobacco_status",
               # original factor columns replaced by pre-encoded integer dummies
               "Gender","Race","Ethnicity","MaritalStatus","GISURH","PriorityGroup","mst_status",
               ".pg_num",".mo")

step_cols_baseline <- c("baseline_k_steps_avg","baseline_k_steps_miss_days")
step_cols_dynamic  <- c("k_steps_7d_avg","k_steps_7d_avg_pctdiff","k_steps_7d_miss_days",
                        "k_steps_7d_all_missing","k_steps_7d_pct_from_baseline","k_steps_7d_sd",
                        "low_activity_days_7d","low_activity_streak",
                        "k_steps_14d_avg","k_steps_14d_sd","k_steps_21d_avg","k_steps_21d_sd")
step_cols <- c(step_cols_baseline, step_cols_dynamic)
static_cols <- setdiff(names(modeling_data),
                       c("Y", drop_cols, step_cols, "days_to_event"))
static_cols <- intersect(static_cols, names(modeling_data))

# ── 70/30 patient-level train/test split ─────────────────────────────────────
set.seed(2026L)
all_pts  <- unique(modeling_data$PatientICN)
train_pts <- sample(all_pts, floor(0.7 * length(all_pts)))
test_pts  <- setdiff(all_pts, train_pts)

keep_all <- c("Y", "PatientICN", "days_to_event", "days_obs_remaining",
              intersect(c(step_cols, static_cols), names(modeling_data)))
modeling_data <- modeling_data %>% select(all_of(intersect(keep_all, names(modeling_data))))

train_data <- modeling_data %>% filter(PatientICN %in% train_pts)
test_data  <- modeling_data %>% filter(PatientICN %in% test_pts)

# ── Save for fast reload ────────────────────────────────────────────────────────────
write_parquet(modeling_data, "data/output/modeling_data.parquet")
write_parquet(tibble(PatientICN = train_pts), "data/output/train_pts.parquet")
write_parquet(tibble(PatientICN = test_pts),  "data/output/test_pts.parquet")
# Pre-compute and save missingness table (week-level and person-level)
miss_tbl_presave <- {
  feat_data  <- modeling_data %>% select(-any_of(c("days_to_event", "Y")))
  feat_cols  <- setdiff(names(feat_data), "PatientICN")

  week_miss <- feat_data %>%
    select(all_of(feat_cols)) %>%
    summarise(across(everything(), ~mean(is.na(.)))) %>%
    pivot_longer(everything(), names_to = "Feature", values_to = "pct_missing_week")

  person_miss <- feat_data %>%
    group_by(PatientICN) %>%
    summarise(across(all_of(feat_cols), ~any(is.na(.))), .groups = "drop") %>%
    select(all_of(feat_cols)) %>%
    summarise(across(everything(), ~mean(.))) %>%
    pivot_longer(everything(), names_to = "Feature", values_to = "pct_missing_person")

  left_join(week_miss, person_miss, by = "Feature") %>%
    mutate(
      pct_missing_week   = round(pct_missing_week   * 100, 1),
      pct_missing_person = round(pct_missing_person * 100, 1)
    ) %>%
    arrange(desc(pct_missing_week))
}
write_parquet(miss_tbl_presave, "data/output/miss_tbl.parquet")
```

```{r missingness-table}
miss_tbl <- arrow::read_parquet("data/output/miss_tbl.parquet") %>%
  mutate(
    label    = coalesce(feat_display_names[Feature], Feature),
    category = classify_feature(Feature)
  ) %>%
  arrange(category, desc(pct_missing_week)) %>%
  select(category, label, pct_missing_week, pct_missing_person)

kbl_miss <- miss_tbl %>%
  gt(groupname_col = "category") %>%
  cols_label(
    label              = "Feature",
    pct_missing_week   = "Weeks Missing (%)",
    pct_missing_person = "Patients Missing (%)"
  ) %>%
  tab_header(title = html(paste0(
    "Table S1. Missingness rates before imputation. ",
    "\u2018Weeks Missing\u2019: % of person-week rows with a missing value. ",
    "\u2018Patients Missing\u2019: % of patients with \u22651 missing week for that feature."
  ))) %>%
  cols_align(align = "left",  columns = label) %>%
  cols_align(align = "right", columns = c(pct_missing_week, pct_missing_person)) %>%
  cols_width(label ~ px(350)) %>%
  tab_options(
    table.font.size       = px(12),
    table.width           = "auto",
    row_group.font.weight = "bold"
  )

kbl_miss
```

```{r model-fitting, eval = FALSE}
# FIT 12 MODELS: 3 feature_sets × 2 horizons × 2 algorithms

fit_m1_one <- function(feature_mode, horizon, model_type,
                       train_data_in = train_data, test_data_in = test_data) {
  message(sprintf("\n[fit_m1_one] %-30s | %2dd | %s", feature_mode, horizon, model_type))
  avail_step   <- intersect(step_cols,   names(train_data_in))
  avail_static <- intersect(static_cols, names(train_data_in))

  keep_feats <- switch(feature_mode,
    model1_steps_only       = avail_step,
    model2_static_only      = avail_static,
    model3_steps_and_static = union(avail_step, avail_static))
  if (length(keep_feats) == 0L) return(NULL)

  model_cols <- c("PatientICN","days_to_event","days_obs_remaining", keep_feats)
  
  # Filter drops rows where all three are true: 
  # * days to event is NA
  # * is not NA days observation remaining
  # * and the window closes before we could observe the event.
  
  # This is not necessary in our current set up because we have more than 30 days of observable data after the last data upload  date but if this is used for more real-time     modeling, this feature will be necessary.
  
  tr <- train_data_in %>% select(all_of(intersect(model_cols, names(train_data_in)))) %>%
    mutate(Y = calc_y_horizon(days_to_event, horizon)) %>%
    # Horizon-specific censoring: exclude no-event rows whose full outcome window is not observable
    filter(!(is.na(days_to_event) & !is.na(days_obs_remaining) & days_obs_remaining < horizon)) %>%
    select(-days_to_event, -days_obs_remaining)   # keep PatientICN for calibration carve + undersampling
  te <- test_data_in  %>% select(all_of(intersect(model_cols, names(test_data_in)))) %>%
    mutate(Y = calc_y_horizon(days_to_event, horizon)) %>%
    filter(!(is.na(days_to_event) & !is.na(days_obs_remaining) & days_obs_remaining < horizon)) %>%
    select(-PatientICN, -days_to_event, -days_obs_remaining)

  # Row-aligned context for stratified sensitivity (identical filter as te; retains PatientICN + miss_days)
  te_context <- test_data_in %>%
    select(PatientICN, any_of("k_steps_7d_miss_days"), days_to_event, days_obs_remaining) %>%
    mutate(Y = calc_y_horizon(days_to_event, horizon)) %>%
    filter(!(is.na(days_to_event) & !is.na(days_obs_remaining) & days_obs_remaining < horizon)) %>%
    select(-days_to_event, -days_obs_remaining)

  # 20% patient-level calibration hold-out (carved BEFORE undersampling)
  # Calibration set retains the true class distribution; undersampling applied to tr_fit only.
  cal_pts    <- sample(unique(tr$PatientICN),
                       size    = max(1L, floor(0.2 * n_distinct(tr$PatientICN))),
                       replace = FALSE)
  tr_fit_raw <- tr %>% filter(!PatientICN %in% cal_pts)   # 80% - model training
  tr_cal_raw <- tr %>% filter( PatientICN %in% cal_pts)   # 20% - Platt calibration only

  # True imbalance ratio computed on model-training portion (before undersampling)
  spw_true <- sum(tr_fit_raw$Y == 0) / max(sum(tr_fit_raw$Y == 1), 1)

  # Logistic: undersample 3:1 (patient-aware) on tr_fit_raw only.
  # XGBoost: full tr_fit_raw + scale_pos_weight = true imbalance ratio.
  tr_under_full <- undersample_by_patient(tr_fit_raw)
  pat_ids_under <- tr_under_full %>% pull(PatientICN)
  tr_under <- as.data.frame(tr_under_full %>% select(-PatientICN))
  tr_full  <- as.data.frame(tr_fit_raw %>% select(-PatientICN))
  
  y_te <- te$Y

  if (model_type == "logistic") {
    message("  fitting logistic regression...")
    tr   <- tr_under
    rec  <- make_recipe_glmnet(tr)
    prp  <- prep(rec, training = tr, retain = TRUE)
    trb  <- bake(prp, new_data = tr); teb  <- bake(prp, new_data = te)
    fit  <- tryCatch(suppressWarnings(glm(Y ~ ., data = trb, family = binomial())),
                     error = function(e) NULL)
    if (is.null(fit)) { message("  SKIP : glm() failed"); return(NULL) }
    message(sprintf("  logistic fit complete  |  %d coefficients", length(coef(fit))))
    pred_raw  <- predict(fit, newdata = teb, type = "response")
    cal_baked <- bake(prp, new_data = as.data.frame(tr_cal_raw %>% select(-PatientICN)))
    pred_cal  <- predict(fit, newdata = cal_baked, type = "response")
    y_cal     <- as.integer(cal_baked$Y)
    coef_tbl <- tryCatch({
      # cluster robust standard errors
      vcov_cl <- sandwich::vcovCL(fit, cluster = data.frame(id = pat_ids_under))
      ct      <- lmtest::coeftest(fit, vcov = vcov_cl)
      as.data.frame(ct) %>%
        tibble::rownames_to_column("term") %>%
        as_tibble() %>%
        rename(estimate  = Estimate,
               std.error = `Std. Error`,
               statistic = `z value`,
               p.value   = `Pr(>|z|)`) %>%
        mutate(odds_ratio = exp(estimate), feature_mode = feature_mode,
               horizon = horizon, model_type = model_type)
    }, error = function(e) {
      warning("vcovCL failed (", conditionMessage(e), "); falling back to standard SEs.")
      broom::tidy(fit) %>%
        mutate(odds_ratio = exp(estimate), feature_mode = feature_mode,
               horizon = horizon, model_type = model_type)
    })
    final_fit <- fit
  } else {
    message(sprintf("  fitting XGBoost (nrounds=200, early_stop=15, spw=%.1f)...", spw_true))
    rec  <- make_recipe_xgboost(tr_full)
    prp  <- prep(rec, training = tr_full, retain = TRUE)
    trb  <- bake(prp, new_data = tr_full); teb  <- bake(prp, new_data = te)
    y_tr <- as.numeric(trb$Y)
    x_tr <- as.matrix(trb %>% select(-Y) %>% mutate(across(everything(), as.numeric)))
    x_te <- as.matrix(teb %>% select(-Y) %>% mutate(across(everything(), as.numeric)))
    # Full training data + scale_pos_weight = true imbalance ratio.
    # Avoids the probability compression that occurs when training on undersampled data.
    dtrain <- xgboost::xgb.DMatrix(x_tr, label = y_tr)
    dtest  <- xgboost::xgb.DMatrix(x_te, label = y_te)
    fit    <- xgboost::xgb.train(
      params = list(objective = "binary:logistic", 
                    eval_metric = "auc",
                    eta = 0.05, # conservative choice
                    max_depth = 5, # models up to 5-way interactions between features
                    min_child_weight = 3, # 3 obs required before split (reduce overfitting)
                    subsample = 0.7, # each tree is trained on 70% random row sample.
                    colsample_bytree = 0.7, # each tree randomly samples 70% of features
                    scale_pos_weight = spw_true), # upweights the gradient contribution of the rare positive class
      data = dtrain,
      nrounds = 200L, 
      verbose = 0L,
      watchlist = list(train = dtrain, test = dtest), early_stopping_rounds = 15L)
    message(sprintf("  XGBoost done  |  best_iteration=%d", fit$best_iteration))
    pred_raw  <- predict(fit, dtest)
    cal_baked <- bake(prp, new_data = as.data.frame(tr_cal_raw %>% select(-PatientICN)))
    x_cal     <- as.matrix(cal_baked %>% select(-Y) %>% mutate(across(everything(), as.numeric)))
    pred_cal  <- predict(fit, xgboost::xgb.DMatrix(x_cal))
    y_cal     <- as.integer(cal_baked$Y)
    imp      <- xgboost::xgb.importance(model = fit)
    coef_tbl <- tibble(term = imp$Feature, gain = imp$Gain,
                       cover = imp$Cover, frequency = imp$Frequency,
                       feature_mode = feature_mode, horizon = horizon, model_type = model_type)
    final_fit <- fit
  }

  # Pre-Platt calibration slope (raw model diagnostic) 
  # Fit logistic regression of the true outcomes on the logit of the raw predictions
  eps_cal       <- 1e-6
  logit_raw     <- log(pmax(eps_cal, pred_raw) / pmax(eps_cal, 1 - pred_raw))
  cal_slope_raw <- tryCatch(
    coef(suppressWarnings(glm(as.integer(y_te) ~ logit_raw, family = binomial())))["logit_raw"],
    error = function(e) NA_real_)

  # Platt scaling - fits a logistic regression on held-out calibration set, apply to test
  # y_cal and pred_cal are set inside the logistic / xgboost branches above.
  pl <- tryCatch(
    suppressWarnings(glm(y_cal ~ pred_cal, family = binomial())),
    error = function(e) NULL)
  pred <- if (!is.null(pl))
    predict(pl, newdata = data.frame(pred_cal = pred_raw), type = "response")
  else pred_raw
  m <- compute_global_metrics(as.integer(y_te), pred)
  message(sprintf("  \u2713 AUC=%.4f  PR-AUC=%.4f  BSS=%.4f  cal_slope=%.3f",
    m$auc, m$pr_auc, m$bss, m$cal_slope))

  list(
    summary = tibble(feature_mode = feature_mode, horizon = horizon, model_type = model_type,
                     auc = round(m$auc, 4), pr_auc = round(m$pr_auc, 4),
                     brier = round(m$brier, 5), bss = round(m$bss, 4),
                     cal_slope_pre_platt  = round(cal_slope_raw, 4),
                     cal_slope_post_platt = round(m$cal_slope, 4),
                     n_test = length(y_te), events_test = sum(as.integer(y_te))),
    roc  = m$roc,
    pred = pred, y = as.integer(y_te),
    preds_df = bind_cols(
      te_context %>% select(PatientICN, any_of("k_steps_7d_miss_days")),
      tibble(y = as.integer(y_te), pred = pred)),
    coef_tbl = coef_tbl,
    calibration = make_calibration_table(pred, as.integer(y_te)),
    fit  = final_fit,
    feature_mode = feature_mode, horizon = horizon, model_type = model_type
  )
}
```

```{r model-run, eval = FALSE}
feature_sets <- c("model1_steps_only","model2_static_only","model3_steps_and_static")
horizons     <- c(14L, 30L)
model_types  <- c("logistic","xgboost")

total_fits <- length(feature_sets) * length(horizons) * length(model_types)
fit_count  <- 0L
m1_results <- list()
for (fm in feature_sets) {
  for (hz in horizons) {
    for (mt in model_types) {
      fit_count <- fit_count + 1L
      message(sprintf("\n══ [%d/%d] ══════════════════════════════════════", fit_count, total_fits))
      res <- tryCatch(
        fit_m1_one(fm, hz, mt),
        error = function(e) { message("  ERROR: ", conditionMessage(e)); NULL })
      if (!is.null(res)) m1_results <- append(m1_results, list(res))
    }
  }
}
message(sprintf("model-run complete: %d/%d fitted successfully.", length(m1_results), total_fits))

m1_summaries  <- bind_rows(lapply(m1_results, `[[`, "summary"))
```

```{r sensitivity-run, eval = FALSE}
# ── Sensitivity B: stratified evaluation by weekly data quality ───────────────
# Uses primary model test-set predictions from m1_results; no model refit.
# Stratifies by k_steps_7d_miss_days:
#   clean  : 0–3 missing days per week (≥4 days with step data)
#   sparse : 4–6 missing days per week (1–3 days with step data)
# Rationale: the primary model is trained on all qualifying weeks (≥1 day of
# step data). Stratified evaluation tests whether the model captures the
# pre-event device-dropout signal present in sparse weeks, rather than relying
# solely on high-quality step counts.

stratum_metrics <- function(preds_sub) {
  if (sum(preds_sub$y) < 5L || nrow(preds_sub) < 10L) return(NULL)
  m <- compute_global_metrics(as.integer(preds_sub$y), preds_sub$pred)
  tibble(
    n_weeks = nrow(preds_sub),
    events  = sum(as.integer(preds_sub$y)),
    auc     = round(m$auc,    4),
    pr_auc  = round(m$pr_auc, 4),
    bss     = round(m$bss,    4)
  )
}

sa_summaries <- bind_rows(lapply(m1_results, function(r) {
  if (is.null(r$preds_df) || !"k_steps_7d_miss_days" %in% names(r$preds_df)) return(NULL)
  preds <- r$preds_df %>%
    mutate(stratum = if_else(k_steps_7d_miss_days <= 3L,
                             "clean (miss \u22643)",
                             "sparse (miss 4\u20136)"))
  strata <- c("clean (miss \u22643)", "sparse (miss 4\u20136)")
  bind_rows(lapply(strata, function(st) {
    sub <- preds %>% filter(stratum == st)
    m   <- stratum_metrics(sub)
    if (is.null(m)) return(NULL)
    bind_cols(
      tibble(feature_mode = r$feature_mode, horizon = r$horizon,
             model_type   = r$model_type,   stratum = st),
      m)
  }))
}))
write_parquet(sa_summaries, "data/output/sa_summaries.parquet")
```

```{r, eval = FALSE}
m1_summaries  <- bind_rows(lapply(m1_results, `[[`, "summary"))

# ── Save full model performance data to xlsx ──────────────────────────────────
writexl::write_xlsx(as.data.frame(m1_summaries),
                    file.path(out_data, "Table2_model_performance_full.xlsx"))

# Wide-format performance table (horizons as columns)
m1_perf_wide <- m1_summaries %>%
  mutate(feature_label = recode(feature_mode, !!!feat_labels),
         algo = dplyr::recode(model_type, xgboost = "XGBoost", logistic = "Logistic")) %>%
  select(feature_label, algo, horizon, auc, pr_auc, bss,
         cal_slope_pre_platt, cal_slope_post_platt, n_test, events_test) %>%
  pivot_wider(
    names_from  = horizon,
    values_from = c(auc, pr_auc, bss, cal_slope_pre_platt, cal_slope_post_platt, n_test, events_test),
    names_glue  = "{.value}_{horizon}d"
  ) %>%
  arrange(algo, feature_label)

write_parquet(m1_perf_wide, "data/output/m1_perf_wide.parquet")

# ── Save test-set predictions with data-quality context ───────────────────────
m1_preds_all <- bind_rows(lapply(m1_results, function(r) {
  if (is.null(r$preds_df)) return(NULL)
  r$preds_df %>% mutate(feature_mode = r$feature_mode,
                        horizon      = r$horizon,
                        model_type   = r$model_type)
}))
write_parquet(m1_preds_all, "data/output/m1_preds_all.parquet")

# ── Save sa_summaries ─────────────────────────────────────────────────────────
write_parquet(sa_summaries, "data/output/sa_summaries.parquet")

# ── Extract and save ROC curve data (30-day) for fast figure rendering ────────
m1_roc_data_30d <- bind_rows(lapply(
  Filter(function(r) r$horizon == 30L, m1_results),
  function(r) data.frame(
    fpr          = 1 - r$roc$specificities,
    tpr          = r$roc$sensitivities,
    feature_mode = r$feature_mode,
    model_type   = r$model_type,
    auc_val      = as.numeric(pROC::auc(r$roc))
  )))
write_parquet(m1_roc_data_30d, "data/output/m1_roc_data_30d.parquet")

# ── Extract and save XGBoost importance tables ────────────────────────────────
m1_xgb_importance <- bind_rows(lapply(
  Filter(function(r) r$model_type == "xgboost", m1_results),
  function(r) as.data.frame(r$coef_tbl) %>%
    mutate(feature_mode = r$feature_mode, horizon = as.integer(r$horizon))))
write_parquet(m1_xgb_importance, "data/output/m1_xgb_importance.parquet")

# ── Extract and save logistic coefficient tables ──────────────────────────────
m1_log_coefs <- bind_rows(lapply(
  Filter(function(r) r$model_type == "logistic", m1_results),
  function(r) as.data.frame(r$coef_tbl)
  # coef_tbl already has feature_mode, horizon, model_type columns
))
write_parquet(m1_log_coefs, "data/output/m1_log_coefs.parquet")

# ── Extract and save calibration data ────────────────────────────────────────
m1_cal_data <- bind_rows(lapply(m1_results, function(r) {
  if (is.null(r$calibration)) return(NULL)
  r$calibration %>%
    mutate(
      feature_mode = r$feature_mode,
      model_type   = r$model_type,
      horizon      = as.integer(r$horizon)
    )
}))
write_parquet(m1_cal_data, "data/output/m1_cal_data.parquet")
```

```{r load-precomputed, include = FALSE}
# ══════════════════════════════════════════════════════════════════════════════
# FAST RELOAD - reads all pre-computed parquet / xlsx outputs.
# Run the eval=FALSE chunks above once to generate these files, then re-render
# with those chunks staying eval=FALSE; this chunk restores every object needed
# by the display sections below.
# ══════════════════════════════════════════════════════════════════════════════
source("cohort_config.R")   # restores max_follow_up_days, max_horizon, inception_cutoff, etc.

# ── Model performance ─────────────────────────────────────────────────────────
m1_summaries <- readxl::read_excel(
  file.path(out_data, "Table2_model_performance_full.xlsx")) %>%
  mutate(horizon = as.integer(horizon))
m1_perf_wide  <- arrow::read_parquet("data/output/m1_perf_wide.parquet") %>%
  mutate(algo = dplyr::recode(algo, Xgboost = "XGBoost", Logistic = "Logistic"))
sa_summaries  <- arrow::read_parquet("data/output/sa_summaries.parquet") %>%
  mutate(horizon = as.integer(horizon))
m1_cal_data   <- arrow::read_parquet("data/output/m1_cal_data.parquet") %>%
  mutate(horizon = as.integer(horizon))

# ── CONSORT counts → individual variables ─────────────────────────────────────
consort_counts <- arrow::read_parquet("data/output/consort_counts.parquet")
list2env(as.list(consort_counts), envir = .GlobalEnv)

# ── Step 5 removal breakdown (event-before-obs-week vs sparse-window) ─────────
.pre  <- arrow::read_parquet("data/output/dt_analysis_pre_step4.parquet") |> as.data.table()
.post <- arrow::open_dataset("data/output/dt_analysis.parquet") |> collect() |> as.data.table()
.removed <- setdiff(unique(.pre$PatientICN), unique(.post$PatientICN))
.rs <- .pre[PatientICN %in% .removed,
  .(first_obs_week_end  = min(weekly_end_date),
    first_event_date    = first(first_event_date),
    censor_date         = first(censor_date),
    obs_window_end      = first(obs_window_end),
    event_within_window = first(event_within_window)),
  by = PatientICN]
n4_removed_early_event   <- .rs[event_within_window == TRUE &
  as.integer(first_obs_week_end - first_event_date) > 0L, .N]
n4_removed_sparse_window <- .rs[as.Date(censor_date) == as.Date(obs_window_end) &
  first_obs_week_end > as.Date(obs_window_end), .N]
rm(.pre, .post, .removed, .rs)

# ── Inline text parameters ────────────────────────────────────────────────────
inline_params <- arrow::read_parquet("data/output/inline_params.parquet")
list2env(as.list(inline_params), envir = .GlobalEnv)

# ── Cohort / analysis data ────────────────────────────────────────────────────
tbl1_data <- arrow::read_parquet("data/output/tbl1_data.parquet") %>%
  mutate(
    first_event_type = factor(first_event_type,
      levels = c("Death","Inpatient","ED Visit","No Event")),
    had_event       = as.integer(first_event_type != "No Event"),
    # Normalize categorical factor levels: dots -> spaces, unconditional title case
    # iconv sanitization ensures valid UTF-8 on VA/Windows before toTitleCase
    across(
      c(Gender, Race, Ethnicity, MaritalStatus, PriorityGroup, mst_status, tobacco_status),
      ~ {
        x <- iconv(gsub(".", " ", as.character(.x), fixed = TRUE), to = "UTF-8", sub = "")
        factor(tools::toTitleCase(tolower(x)))
      }
    ),
    # GISURH: explicit letter-code expansion; toupper() handles mixed-case input
    GISURH = relevel(factor(dplyr::recode(toupper(trimws(as.character(GISURH))),
      "H" = "Highly Rural",
      "I" = "Insular",
      "R" = "Rural",
      "U" = "Urban",
      .default = "Unknown"
    )), ref = "Urban"),
    MaritalStatus = relevel(MaritalStatus, ref = "Married"),
    py_ed           = factor(py_ed),
    py_urgent       = factor(py_urgent),
    py_appt_no_show = factor(py_appt_no_show)
  )
dt_analysis <- arrow::open_dataset("data/output/dt_analysis.parquet") %>%
  collect() %>% as.data.table()

# ── Model component data for figures ─────────────────────────────────────────
m1_roc_data_30d   <- arrow::read_parquet("data/output/m1_roc_data_30d.parquet")
m1_xgb_importance <- arrow::read_parquet("data/output/m1_xgb_importance.parquet")
m1_log_coefs      <- arrow::read_parquet("data/output/m1_log_coefs.parquet")
```

---

# Methods

## Data

This study uses Patient-Generated Health Data (PGHD) from the VA Wearable Technology Program, in which Veterans voluntarily enrolled and shared wearable device data (Fitbit, Garmin, or Apple Health). Data were obtained from **`r data_start_date`** through **`r data_end_date`**. Daily step counts from Fitbit and Garmin devices were prioritized (using the higher value when both were present); Apple Health data served as an alternative when Fitbit/Garmin data were unavailable. A total of `r format(n_pdays_with_steps, big.mark = ",")` patient-days had usable step data; Fitbit/Garmin data were available for `r pct_fitgarmin`% and Apple Health for the remainder. Daily step counts of zero were treated as missing, and outliers were removed using the 1.5×IQR method applied to raw device columns. Where step data were missing but exercise duration (minutes) was recorded, steps were imputed from a linear regression model fit on matched days (`steps ~ exercise_minutes`).

## Cohort Derivation (Figure 1)

The analytic cohort was drawn from Veterans who had enrolled in the VA Wearable Technology Program with PGHD data available from **`r data_start_date`** through **`r data_end_date`**. The enrollment criterion was a first PGHD upload between **`r data_start_date`** and **`r followup_cutoff_label`** (N = `r format(n1_pts, big.mark=",")`). The lower bound marks the program inception date; the upper bound ensures at least `r max_follow_up_days %/% 7L` weeks of potential observation remain before the data extract end date, providing a uniform minimum observation window for every included patient. We retained all person-weeks with at least one day of step data (device-inactive weeks where no steps were uploaded are excluded implicitly by the rolling-window computation). We retained only patients who had at least one qualifying week in the baseline period (calendar weeks 1–6; days 1–42 after each patient's actual first upload date) and at least one qualifying observation week (week 7 onward). Calendar weeks are indexed relative to each patient's actual first upload date (week 1 = days 1–7, week 2 = days 8–14, etc.). Patients whose composite event (ED visit, inpatient admission, or death) occurred within the first 42 days after first upload (calendar weeks 1–6) were additionally excluded, as no event-free anchor period exists for them. Observation windows were then censored at the date of a first composite event or at `r max_follow_up_days %/% 7L` weeks from first upload, whichever came first. The final analytic cohort included **`r format(n4_pts, big.mark=",")` Veterans** contributing **`r format(n4_wks, big.mark=",")` person-weeks** of observation (weeks 7+).

## Step Feature Construction

At each observation week, a suite of step-derived features was computed from the 7-day, 14-day, and 21-day rolling windows anchored to the calendar week end-date. The primary current-activity features include the 7-day rolling average daily step count (`k_steps_7d_avg`, thousands), the rolling standard deviation (`k_steps_7d_sd`), the number of missing days in the window (`k_steps_7d_miss_days`), and the count of days with fewer than 1,000 steps (`low_activity_days_7d`). Two change metrics capture temporal trends: percent change from the average of the prior three 7-day windows (`k_steps_7d_avg_pctdiff`, winsorized at plus or minus 500%) and percent change from the 6-week baseline average (`k_steps_7d_pct_from_baseline`). Analogous 14-day and 21-day rolling averages and standard deviations provide longer-horizon smoothed signals. Two baseline summary features (mean steps and mean missing days across weeks 1-6) anchor the within-person normalization.

Taken together, the step features implement a Mundlak-style decomposition for a signal for high or lower activity relative to normal activity. `baseline_k_steps_avg`, the patient's mean daily step count across the six baseline weeks, fixed at enrollment, serves as the between-person anchor, capturing stable individual-level activity capacity. `k_steps_7d_pct_from_baseline` and `k_steps_7d_avg_pctdiff` represent within-person deviations from that anchor, capturing departures from the patient's own history over medium and short horizons respectively. Because the baseline is computed exclusively from pre-observation data (weeks 1–6), all deviation features are computable at any future prediction point without requiring knowledge of future step counts, a prerequisite for prospective deployment.

## Outcome Definition

The primary outcome was a composite of ED visit, inpatient admission, or death occurring within the prediction horizon ($h$ = 14 or 30 days) following a given observation week's end date. The outcome was labeled positive ($Y = 1$) if any component occurred within 1 to $h$ days (`days_to_event` $\in [1, h]$), and negative ($Y = 0$) otherwise. For each prediction horizon $h$, observation weeks for non-event patients were excluded if fewer than $h$ days remained before the end of the patient's observation window, as the outcome classification could not be confirmed over the full horizon. This ensures that 14-day models retain more person-weeks than 30-day models for the same patients.

## Statistical Analysis

We fit three feature sets: (1) step-derived features only (baseline + dynamic, *n* = 14 features); (2) static clinical features only (demographics, comorbidity flags, prior healthcare utilization, vitals, labs, and social history); and (3) the combined set. For each feature set and each prediction horizon (14-day and 30-day), we trained two algorithm types: penalized logistic regression (with elastic-net preprocessing) and gradient-boosted trees (XGBoost). Patients were randomly assigned to training (70%) and testing (30%) sets; all observation-weeks for a given patient remained in the same partition. To address class imbalance, logistic regression models were trained on patient-level 3:1 undersampled data; XGBoost models were trained on the full (non-undersampled) training data with `scale_pos_weight` set to the true negative-to-positive ratio, avoiding the probability compression that results from training and scoring on different class distributions.

To ensure accurate absolute risk estimates, predicted probabilities were calibrated using Platt scaling (logistic regression of observed outcomes on raw model log-odds). Calibration models were fit on a held-out 20% of training patients (patient-level split, preserving the original class distribution) and applied to the independent test set, ensuring no test-set information was used to fit the calibration model. All discrimination and calibration metrics were computed on the Platt-calibrated test-set probabilities. Model performance was evaluated using the Area Under the Receiver Operating Characteristic Curve (AUC-ROC), the Area Under the Precision-Recall Curve (PR-AUC), the post-calibration slope (target: 1.0), and the Brier Skill Score (BSS = 1 $-$ Brier/Brier$_{\text{null}}$, where Brier$_{\text{null}}$ = $\bar{p}(1-\bar{p})$). BSS accounts for the low baseline event rate (<3% per week): BSS = 0 indicates no improvement over a null model predicting the marginal event rate, and BSS = 1 indicates perfect probabilistic skill. The pre-calibration slope is additionally reported as a diagnostic of raw model output compression.

To evaluate whether the primary model captures predictive signal from weeks with sparse device engagement, test-set predictions were stratified post-hoc by weekly data quality: "clean" weeks (0–3 missing days; greater than 4 days with step data) and "sparse" weeks (4–6 missing days; 1–3 days with step data). Discrimination and skill metrics (AUC, PR-AUC, BSS) were computed separately for each stratum without model refit, directly testing whether weeks with low device engagement (which may reflect pre-event health deterioration) are predicted as effectively as high-quality weeks.

## Figure 1. Cohort Derivation

```{r consort-table}
consort_df <- tribble(
  ~Step, ~Criterion, ~Veterans, ~`Person-weeks`,
  "1. Enrolled PGHD cohort",
    paste0("Veterans with first PGHD upload between ", data_start_date, " and ", followup_cutoff_label,
           ". Person-weeks with \u22651 day of step data are retained."),
    format(n1_pts, big.mark = ","), format(n1_wks, big.mark = ","),
  "2. Baseline and observation weeks present",
    "At least one qualifying week in baseline (calendar weeks 1\u20136, days 1\u201342 after first upload) and at least one qualifying observation week (week 7+)",
    format(n3_pts, big.mark = ","), format(n3_wks, big.mark = ","),
  "3. Event-free baseline required",
    "Exclude patients whose composite event occurred within the first 42 days after first upload (calendar weeks 1\u20136)",
    format(n3b_pts, big.mark = ","), format(n3b_wks, big.mark = ","),
  "4. Censor at first event or follow-up window end",
    paste0(
      "Remove person-weeks after first composite event or ", max_follow_up_days %/% 7L,
      "-week window end. Of the ", format(n4_before - n4_pts, big.mark = ","),
      " patients excluded: (",
      "a) ", format(n4_removed_early_event, big.mark = ","),
      " had their composite event before their first qualifying observation week end-date",
      " (event occurred during or before the week-7 transition period, leaving no complete prediction window); ",
      "(b) ", format(n4_removed_sparse_window, big.mark = ","),
      " enrolled early but had insufficient PGHD data quality during their ",
      max_follow_up_days %/% 7L,
      "-week observation window - their first qualifying observation week only appeared after the window closed,",
      " reflecting prolonged gaps or low upload engagement rather than a clinical event."),
  format(n4_pts, big.mark = ","),
  format(n4_wks, big.mark = ",")
)
```
```{r}
consort_df %>%
  gt() %>%
  tab_header(title = html("Figure 1. Cohort derivation flow table for Study 1: Person-weeks shown for observation period (weeks 7+) at steps 3\u20134; earlier rows show total qualifying weekly rows before baseline/observation split.")) %>%
  cols_align(align = "left",  columns = c(Step, Criterion)) %>%
  cols_align(align = "right", columns = c(Veterans, `Person-weeks`)) %>%
  cols_width(
    Step           ~ px(288),
    Criterion      ~ px(480),
    Veterans       ~ px(112),
    `Person-weeks` ~ px(112)
  ) %>%
  tab_options(table.font.size = px(12), table.width = "auto")
```

---

## Cohort Derivation Diagnostics

The sections below inspect each CONSORT exclusion step to verify that exclusions are appropriate. These are computed by re-applying each filter in isolation from a snapshot of the weekly data (`dt_weekly_snapshot`) taken after rolling-feature construction but before any CONSORT filtering.

```{r consort-diagnostics, include = FALSE}
# CONSORT DIAGNOSTICS - re-runs each step in isolation to expose who is
# excluded and why.  Loads from snapshots saved in consort-pipeline.
# Results saved to data/output/consort_diag_step*.parquet for display chunks.

dt_wk  <- open_dataset("data/output/dt_weekly_snapshot.parquet") %>% collect() %>% as.data.table()

# Distribution of first_upload_date for annotation

step1_dist <- dt_wk %>%
  as_tibble() %>%
  distinct(PatientICN, .keep_all = TRUE) %>%
  transmute(first_upload_date = as.Date(first_upload_date)) %>%
  dplyr::count(month = lubridate::floor_date(first_upload_date, "month"), name = "n_patients")

step1_dist   %>% write_parquet("data/output/consort_diag_step1_dist.parquet")
rm(step1_dist)

# 
# Step 2 Sensitivity: weekly data quality (miss_days threshold)
# Thresholds: require ≥1..7 days present → miss_days ≤ 6,5,4,3,2,1,0
# 
```

```{r, include = FALSE}
# Apply step-1 primary first so counts are comparable to primary analysis
dt_s1 <- dt_wk[first_upload_date < as.Date("2025-03-17")]

step2_thresholds <- 0:6   # miss_days ≤ this value
step2_min_days   <- 7 - step2_thresholds  # days present required

step2_diag <- map2_dfr(step2_thresholds, step2_min_days, function(thresh, min_d) {
  kept <- dt_s1[k_steps_7d_miss_days <= thresh]
  tibble(
    miss_days_threshold = thresh,
    days_required       = min_d,
    label               = paste0("≤", thresh, " missing (≥", min_d, " days present)",
                                 if_else(thresh == 6, " [current]", "")),
    n_patients          = uniqueN(kept$PatientICN),
    n_weeks             = nrow(kept),
    pct_patients_retained = round(100 * uniqueN(kept$PatientICN) / uniqueN(dt_s1$PatientICN), 1),
    pct_weeks_retained    = round(100 * nrow(kept) / nrow(dt_s1), 1)
  )
})

# Distribution of miss_days across all rows (after step-1)
step2_dist <- dt_s1 %>%
  as_tibble() %>%
  dplyr::count(miss_days = k_steps_7d_miss_days, name = "n_weeks") %>%
  mutate(pct = round(100 * n_weeks / sum(n_weeks), 1))

step2_diag %>% write_parquet("data/output/consort_diag_step2.parquet")
step2_dist %>% write_parquet("data/output/consort_diag_step2_dist.parquet")
rm(step2_diag, step2_dist, dt_s1)
```
```{r, include = FALSE}
# Step 3 Sensitivity: baseline length (2,4,6,8,10,12 weeks)

dt_s2 <- dt_wk[first_upload_date < as.Date("2025-03-16") & k_steps_7d_miss_days <= 3]

step3_baselines <- c(2L, 4L, 6L, 8L, 10L, 12L)

step3_diag <- map_dfr(step3_baselines, function(bl) {
  min_total <- bl + 1L   # bl baseline + ≥1 observation week
  pt_wk     <- dt_s2[, .(tot = .N), by = PatientICN]
  kept_pts  <- pt_wk[tot >= min_total, PatientICN]
  kept      <- dt_s2[PatientICN %in% kept_pts]
  # Observation weeks (post-baseline)
  obs_wks   <- kept[weekly_interval_index > bl]
  tibble(
    baseline_wks      = bl,
    label             = paste0(bl, " weeks", if_else(bl == 6L, " [current]", "")),
    n_patients        = uniqueN(kept$PatientICN),
    n_obs_weeks       = nrow(obs_wks),
    pct_of_s2_pts     = round(100 * uniqueN(kept$PatientICN) / uniqueN(dt_s2$PatientICN), 1)
  )
})

# Total weeks distribution (to show natural cliff)
step3_dist <- dt_s2 %>%
  as_tibble() %>%
  group_by(PatientICN) %>%
  summarise(total_wks = n(), .groups = "drop") %>%
  dplyr::count(total_wks, name = "n_patients")

step3_diag %>% write_parquet("data/output/consort_diag_step3.parquet")
step3_dist %>% write_parquet("data/output/consort_diag_step3_dist.parquet")
rm(step3_diag, step3_dist, dt_s2)
```
```{r, include = FALSE}
# Step 4 Diagnostic: understand week-removal at censoring step

pre4 <- open_dataset("data/output/dt_analysis_pre_step4.parquet") %>% collect() %>% as.data.table()
pre4[, weekly_end_date    := as.Date(weekly_end_date)]
pre4[, first_event_date   := as.Date(first_event_date)]
pre4[, first_upload_date  := as.Date(first_upload_date)]
pre4[, obs_window_end := as.Date(first_upload_date) + max_follow_up_days]

pre4[, event_within_window := !is.na(first_event_date) & first_event_date <= obs_window_end + max_horizon]
pre4[, censor_date := fifelse(
  event_within_window,
  as.character(first_event_date),
  as.character(obs_window_end))]
pre4[, censor_date := as.Date(censor_date)]
pre4[, keep_row := weekly_end_date <= censor_date]

# Summary: patients before vs after (should be same or very close)
step4_patient_summary <- tibble(
  stage          = c("Pre-step 5 (post-step 4 obs weeks)", "Post-step 5 (censored)"),
  n_patients     = c(uniqueN(pre4$PatientICN),
                     uniqueN(pre4[keep_row == TRUE]$PatientICN)),
  n_weeks        = c(nrow(pre4), nrow(pre4[keep_row == TRUE])),
  n_weeks_removed = c(NA_integer_, nrow(pre4[keep_row == FALSE]))
)

# Reason for row removal
step4_removal_reason <- pre4[keep_row == FALSE] %>%
  as_tibble() %>%
  mutate(
    removal_reason = case_when(
      event_within_window ~ "Event censoring (row after first event)",
      TRUE                ~ paste0(max_follow_up_days %/% 7L, "-week / end-of-window censoring")
    )
  ) %>%
  dplyr::count(removal_reason, name = "n_weeks_removed")

# Weekly_interval_index distribution for removed rows
step4_removed_index_dist <- pre4[keep_row == FALSE] %>%
  as_tibble() %>%
  dplyr::count(weekly_interval_index, name = "n_removed") %>%
  arrange(weekly_interval_index)

# Distribution of days_to_event for event-censored rows
step4_event_days_dist <- pre4[keep_row == FALSE & event_within_window == TRUE] %>%
  as_tibble() %>%
  transmute(
    days_to_event_at_removal = as.integer(first_event_date - weekly_end_date)
  )

step4_patient_summary    %>% write_parquet("data/output/consort_diag_step4_summary.parquet")
step4_removal_reason     %>% write_parquet("data/output/consort_diag_step4_reason.parquet")
step4_removed_index_dist %>% write_parquet("data/output/consort_diag_step4_index.parquet")
step4_event_days_dist    %>% write_parquet("data/output/consort_diag_step4_event_days.parquet")
rm(pre4, step4_patient_summary, step4_removal_reason, step4_removed_index_dist, step4_event_days_dist)
```

### Step 1: Distribution of first PGHD upload date by month

```{r diag-step1-chart, echo=FALSE, fig.height=3.5, fig.cap="Distribution of first PGHD upload date by month. Vertical lines mark the three cutoff options. Patients to the right of a cutoff are excluded at Step 1."}
if (file.exists("data/output/consort_diag_step1_dist.parquet")) {
  d1_dist <- open_dataset("data/output/consort_diag_step1_dist.parquet") %>% collect()
  # column is 'n_patients' if regenerated, or 'n' from older runs - handle both
  if (!"n_patients" %in% names(d1_dist) && "n" %in% names(d1_dist))
    d1_dist <- d1_dist %>% rename(n_patients = n)
  cutoff_lines <- tibble(
    cutoff = as.Date(c("2024-09-16","2025-03-16","2025-09-16")),
    label  = c("-6 mo","Current","+ 6 mo")
  )
  ggplot(d1_dist, aes(x = month, y = n_patients)) +
    geom_col(fill = "#4575b4", alpha = 0.75) +
    geom_vline(data = cutoff_lines, aes(xintercept = as.numeric(cutoff), linetype = label),
               color = c("#d73027","#1a1a1a","#69b3a2"), linewidth = 0.7) +
    scale_linetype_manual(values = c("Current"="solid","-6 mo"="dashed","+ 6 mo"="dotdash"),
                          name = "Cutoff") +
    labs(x = "First upload month", y = "Patients",
         title = "Step 1: Distribution of first PGHD upload date") +
    theme_science() +
    theme(legend.position = "right", legend.text = element_text(size = 8))
}
```

---

### Step 2: Weekly Data Quality Threshold Sensitivity

```{r diag-step2, echo=FALSE}
if (file.exists("data/output/consort_diag_step2.parquet")) {
  d2      <- open_dataset("data/output/consort_diag_step2.parquet") %>% collect()
  d2_dist <- open_dataset("data/output/consort_diag_step2_dist.parquet") %>% collect()

  d2 %>%
    select(-miss_days_threshold, -days_required) %>%
    setNames(c("Quality criterion", "Patients", "Person-weeks",
               "% patients retained", "% weeks retained")) %>%
    gt() %>%
    tab_header(title = "Step 2 Sensitivity: Effect of varying the required minimum upload days per 7-day window.") %>%
    cols_align(align = "left",  columns = `Quality criterion`) %>%
    cols_align(align = "right", columns = c(Patients, `Person-weeks`, `% patients retained`, `% weeks retained`)) %>%
    tab_options(table.font.size = px(12), table.width = "auto")
} else {
  cat("*Step 2 diagnostics not yet computed - run the `consort-diagnostics` chunk.*")
}
```

```{r diag-step2-chart, echo=FALSE, fig.height=3.5, fig.cap="Distribution of missing days per 7-day window across all qualifying person-weeks (after Step 1). The current threshold retains miss_days ≤ 6) retains."}
if (file.exists("data/output/consort_diag_step2_dist.parquet")) {
  d2_dist <- open_dataset("data/output/consort_diag_step2_dist.parquet") %>% collect()
  ggplot(d2_dist %>% filter(!is.na(miss_days)),
         aes(x = factor(miss_days), y = n_weeks)) +
    geom_col() +
    geom_text(aes(label = paste0(pct, "%")), vjust = -0.3, size = 2.8) +
    labs(x = "Missing days in 7-day window", y = "Person-weeks",
         title = "Step 2: Distribution of missing days per week") +
    scale_y_continuous() +
    theme_science() +
    theme(legend.position = "top")
}
```

---

### Step 3: Minimum Baseline Length Sensitivity

```{r diag-step3, echo=FALSE}
  d3      <- open_dataset("data/output/consort_diag_step3.parquet") %>% collect()
  d3_dist <- open_dataset("data/output/consort_diag_step3_dist.parquet") %>% collect()

  d3 %>%
    select(-baseline_wks) %>%
    setNames(c("Baseline length", "Patients retained", "Observation person-weeks",
               "% of post-step2 patients")) %>%
    gt() %>%
    tab_header(title = "Step 3 Sensitivity: Effect of varying the minimum required baseline length.") %>%
    cols_align(align = "left",  columns = `Baseline length`) %>%
    cols_align(align = "right", columns = c(`Patients retained`, `Observation person-weeks`, `% of post-step2 patients`)) %>%
    tab_options(table.font.size = px(12), table.width = "auto")
```

```{r diag-step3-chart, echo=FALSE, fig.height=3.5, fig.cap="Cumulative distribution of total qualifying weeks per patient (after Steps 1–2). Patients to the left of each vertical line would be excluded by the corresponding minimum baseline requirement."}
if (file.exists("data/output/consort_diag_step3_dist.parquet")) {
  d3_dist <- open_dataset("data/output/consort_diag_step3_dist.parquet") %>% collect()
  # Cumulative distribution
  d3_cum <- d3_dist %>%
    arrange(total_wks) %>%
    mutate(cum_pct = 100 * cumsum(n_patients) / sum(n_patients))
  bl_lines <- tibble(
    x     = c(3, 5, 7, 9, 11, 13),   # min total = baseline + 1
    label = paste0(c(2,4,6,8,10,12), "wk BL")
  )
  ggplot(d3_cum, aes(x = total_wks, y = cum_pct)) +
    geom_step(color = "#4575b4", linewidth = 0.8) +
    geom_vline(data = bl_lines, aes(xintercept = x, linetype = label),
               color = "grey50", linewidth = 0.5) +
    scale_linetype_manual(values = rep(c("dashed","dotdash"), 3), name = "Min total wks") +
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    labs(x = "Total qualifying weeks per patient", y = "Cumulative % of patients",
         title = "Step 3: Cumulative distribution of total qualifying weeks") +
    theme_science() +
    theme(legend.position = "right", legend.text = element_text(size = 7.5))
}
```

---


```{r diag-step5-charts, echo=FALSE, fig.height=4.5, fig.cap="Left: Distribution of weekly_interval_index for removed person-weeks (weeks dropped by censoring). Right: Distribution of days_to_event at the time of removal for event-censored rows (i.e., how far in the future was the event relative to the dropped week?)."}
if (all(file.exists(c("data/output/consort_diag_step4_index.parquet",
                       "data/output/consort_diag_step4_event_days.parquet")))) {
  d4_idx   <- open_dataset("data/output/consort_diag_step4_index.parquet")      %>% collect()
  d4_evtd  <- open_dataset("data/output/consort_diag_step4_event_days.parquet") %>% collect()

  p_idx <- ggplot(d4_idx, aes(x = weekly_interval_index, y = n_removed)) +
    geom_col(fill = "#d73027", alpha = 0.8) +
    labs(x = "Week index (since first upload)", y = "Weeks removed",
         title = "Removed weeks by week index") +
    theme_science()

  p_evtd <- ggplot(d4_evtd %>% filter(!is.na(days_to_event_at_removal)),
                   aes(x = days_to_event_at_removal)) +
    geom_histogram(binwidth = 7, fill = "#4575b4", alpha = 0.8, color = "white") +
    labs(x = "Days from dropped week to first event", y = "Person-weeks",
         title = "Event-censored: proximity to event") +
    theme_science()

  p_idx + p_evtd
}
```

---

# Results: Study 1 - Pre-Event PGHD Surveillance Cohort

## Table 1. Sample Characteristics

```{r table1, echo=FALSE}
# Convert k_steps columns to full step counts for Table 1 display 
tbl1_data <- tbl1_data %>%
  mutate(
    baseline_steps_avg  = baseline_k_steps_avg  * 1000,
    obs_steps_avg       = obs_k_steps_avg        * 1000
  )

# ── % of observation weeks with qualifying data coverage ──────────────────────
# Numerator: qualifying (miss_days ≤ 3) weeks in obs period (weeks 7+) = rows in dt_analysis
# Denominator: total 7-day windows from obs start (first_upload + 42d) to censor date
.obs_coverage <- as.data.table(dt_analysis)[
  , .(qualifying_weeks  = .N,
      obs_start         = first(first_upload_date) + 42L,
      censor_d          = max(weekly_end_date)),
  by = PatientICN
][
  , possible_weeks := pmax(1L, as.integer(censor_d - obs_start) %/% 7L + 1L)
][
  , pct_obs_coverage := round(100 * qualifying_weeks / possible_weeks, 1)
]
tbl1_data <- tbl1_data %>%
  left_join(as_tibble(.obs_coverage[, .(PatientICN, pct_obs_coverage)]),
            by = "PatientICN")
rm(.obs_coverage)

# ── Save Table 1 underlying data to xlsx ──────────────────────────────────────
writexl::write_xlsx(as.data.frame(tbl1_data),
                    file.path(out_data, "Table1_cohort_characteristics.xlsx"))

# Determine which py2_ columns are actually present
py2_cols <- names(tbl1_data)[str_detect(names(tbl1_data), "^py2_")]

# ── Explicit variable groups (controls ordering and group headers) ─────────────
grp_pghd <- intersect(
  c("baseline_steps_avg", "baseline_k_steps_miss",
    "obs_steps_avg", "obs_miss_days_avg", "total_weeks", "pct_obs_coverage"),
  names(tbl1_data))

grp_can <- intersect(
  c("Age", "Gender", "Race", "Ethnicity", "MaritalStatus", "GISURH", "PriorityGroup", "season",
    py2_cols, "cci_score",
    "py_ed", "py_urgent", "py_appt_no_show",
    "bmi_fitted_mean", "map_fitted_mean", "recent_pulse",
    "bun_value_mean", "albumin_value_mean", "leuk_value_mean",
    "mst_status", "current_smoker", "never_smoker"),
  names(tbl1_data))

grp_events <- intersect(
  c("had_event", "days_to_event_from_start"),
  names(tbl1_data))

tbl1_vars <- c(grp_pghd, grp_can, grp_events)

# Continuous py2_ variables - kept numeric, not coerced to factor
py2_continuous <- c("py2_mwi_score", "py2_IVC_los", "py2_IVC_n")

py2_label_map <- c(
  py2_acute_renal_failure      = "Acute renal failure (2yr)",
  py2_artificial_openings      = "Artificial opening / ostomy (2yr)",
  py2_chronic_pancreatitis     = "Chronic pancreatitis (2yr)",
  py2_chronic_ulcer_skin       = "Chronic skin ulcer (2yr)",
  py2_congestive_heart_failure = "Congestive heart failure (2yr)",
  py2_drug_alcohol_dependence  = "Substance use disorder (2yr)",
  py2_drug_alcohol_psychosis   = "Substance-induced psychosis (2yr)",
  py2_lung_severe_cancers      = "Lung / severe cancer (2yr)",
  py2_metastatic_cancer        = "Metastatic cancer (2yr)",
  py2_pregnancy                = "Pregnancy (2yr)",
  py2_heart_arrhythmias        = "Cardiac arrhythmia (2yr)",
  py2_financial_sdoh           = "Financial hardship / SDOH (2yr)",
  py2_housing_instability      = "Housing instability (2yr)",
  py2_nonspecific_psychosocial = "Nonspecific psychosocial problem (2yr)",
  py2_dialysis_access          = "Dialysis access procedure (2yr)",
  py2_dialysis_status          = "Dialysis status (2yr)",
  py2_mwi_score                = "Multimorbidity Weighted Index score (2yr)",
  py2_IVC_n                    = "Community care stays, count (among admitted, 2yr)",
  py2_IVC_los                  = "Community care, total LOS days (among admitted, 2yr)",
  py2_IVC                      = "Community care stays, any (2yr)",

  py2_inpat_any                = "Any inpatient admission (2yr)",
  py2_inpat_med_surg           = "Med/surgical admission (2yr)",
  py2_inpat_mental_health      = "Mental health admission (2yr)",
  py2_inpat_nursing_home       = "Nursing home admission (2yr)",
  py2_inpat_other              = "Other inpatient admission (2yr)"
)
base_label_map <- list(
  baseline_steps_avg       = "Baseline avg. steps (steps/day)",
  baseline_k_steps_miss    = "Baseline avg. missing days/week",
  obs_steps_avg            = "Observation avg. steps (steps/day)",
  obs_miss_days_avg        = "Observation avg. missing days/week",
  total_weeks              = "Total observation weeks (post-baseline)",
  Age                      = "Age (years)",
  Gender                   = "Sex",
  Race                     = "Race",
  Ethnicity                = "Ethnicity",
  MaritalStatus            = "Marital status",
  GISURH                   = "Rural/Urban (GISURH)",
  PriorityGroup            = "VA Priority Group",
  season                   = "Season of observation",
  py_ed                    = "Any ED visit (prior year)",
  py_urgent                = "Any urgent care (prior year)",
  py_appt_no_show          = "Appointment no-show (prior year)",
  bmi_fitted_mean          = "BMI (spline-fitted mean)",
  map_fitted_mean          = "MAP (spline-fitted mean, mmHg)",
  recent_pulse             = "Recent pulse (bpm)",
  bun_value_mean           = "BUN (mean, mg/dL)",
  albumin_value_mean       = "Albumin (mean, g/dL)",
  leuk_value_mean          = "Leukocytes (mean, 10^3/\u00b5L)",
  cci_score                = "Charlson Comorbidity Index (CCI) score",
  mst_status               = "Military sexual trauma",
  current_smoker           = "Current smoker",
  never_smoker             = "Never smoker",
  had_event                = "Had composite event",
  days_to_event_from_start = "Days from first upload to event",
  pct_obs_coverage         = "Observation weeks with data coverage (%)"
)
all_labels    <- c(as.list(py2_label_map), base_label_map)
present_labels <- all_labels[intersect(names(all_labels), tbl1_vars)]

# Mark reference groups & convert NA → "Unknown" for categorical display 
mark_ref <- function(x, ref_lv = NULL) {
  # Strip any prior "(ref)" tags so the function is idempotent across re-runs
  x_chr <- gsub(" \\(ref\\)$", "", as.character(x))
  x <- factor(x_chr)
  if (is.null(ref_lv)) ref_lv <- levels(x)[1]
  new_lv <- c(paste0(ref_lv, " (ref)"), setdiff(levels(x), c(ref_lv, "Unknown")), "Unknown")
  x_chr[is.na(x_chr)] <- "Unknown"
  x_chr[x_chr == ref_lv] <- paste0(ref_lv, " (ref)")
  factor(x_chr, levels = new_lv)
}
tbl1_data <- tbl1_data %>%
  mutate(
    Gender         = mark_ref(Gender),
    Race           = mark_ref(Race, "White"),
    Ethnicity      = mark_ref(Ethnicity, "Not Hispanic or Latino"),
    MaritalStatus  = mark_ref(MaritalStatus, "Married"),
    GISURH         = mark_ref(GISURH, "Urban"),
    PriorityGroup  = mark_ref(PriorityGroup, "Group 1"),
    mst_status     = mark_ref(mst_status),
    tobacco_status = mark_ref(tobacco_status)
  )

tbl_obj <- tbl1_data %>%
  select(first_event_type, all_of(tbl1_vars)) %>%
  tbl_summary(
    by      = first_event_type,
    missing = "ifany",
    statistic = list(
      all_continuous()  ~ "{mean} ({sd}); {median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(all_continuous() ~ 1),
    label  = present_labels
  )

# Add visual section headers - requires gtsummary >= 2.0; falls back gracefully
tbl_obj <- tryCatch(
  tbl_obj %>%
    add_variable_group_header(
      list(
        "PGHD / Step Features"               = grp_pghd,
        "CAN-Score Analog Clinical Features"  = grp_can,
        "Event Variables"                     = grp_events
      )
    ),
  error = function(e) tbl_obj
)

tbl1_display <- tbl_obj %>%
  add_overall(last = FALSE) %>%
  add_p(pvalue_fun = ~style_pvalue(.x, digits = 2)) %>%
  bold_labels() %>%
  modify_caption(
    "**Table 1. Sample characteristics by first event type.** Variables grouped as: (1) PGHD / Step Features, (2) CAN-Score Analog Clinical Features, (3) Event Variables. Continuous: mean (SD); median (IQR); categorical: N (%). p-values from Kruskal-Wallis (continuous) or \u03c7\u00b2 (categorical) tests."
  )
tbl1_display
```

```{r table1-download, echo=FALSE}
if (requireNamespace("downloadthis", quietly = TRUE)) {
  gtsummary::as_tibble(tbl1_display) %>%
    downloadthis::download_this(
      output_name      = "Table1_cohort_characteristics",
      output_extension = ".xlsx",
      button_label     = "Download Table 1 data (xlsx)",
      button_type      = "default",
      class            = "btn-sm"
    )
}
```

---

## Figure 2. Time to First Composite Event

```{r fig2-survival, echo=FALSE, fig.height=4.5, fig.cap="Figure 2. Kaplan-Meier survival curve for time from first PGHD upload to first composite event (ED visit, inpatient admission, or death) among the analytic cohort. The y-axis shows the proportion of patients remaining event-free. Tick marks indicate censored observations. Shaded band represents the 95% pointwise confidence interval."}

if (!requireNamespace("survival", quietly = TRUE))
  try(install.packages("survival", repos = "https://cloud.r-project.org", quiet = TRUE), silent = TRUE)
library(survival)

km_data <- dt_analysis[, .(
  first_upload_date = first(first_upload_date),
  first_event_date  = first(first_event_date),
  first_event_type  = first(first_event_type),
  last_obs_date     = max(weekly_end_date)
), by = PatientICN] %>%
  as_tibble() %>%
  mutate(
    surv_time  = as.numeric(if_else(
      !is.na(first_event_date),
      difftime(first_event_date, first_upload_date, units = "days"),
      difftime(last_obs_date,    first_upload_date, units = "days"))),
    event_flag = as.integer(!is.na(first_event_date))
  ) %>%
  filter(surv_time >= 0)

km_fit  <- survfit(Surv(surv_time, event_flag) ~ 1, data = km_data)
km_df   <- data.frame(
  time   = km_fit$time,
  surv   = km_fit$surv,
  lower  = km_fit$lower,
  upper  = km_fit$upper,
  n.cens = km_fit$n.censor
)

n_events  <- sum(km_data$event_flag)
n_total   <- nrow(km_data)
med_surv  <- summary(km_fit)$table["median"]
surv_180  <- round(100 * summary(km_fit, times = 180)$surv,  1)
surv_365  <- round(100 * summary(km_fit, times = 365)$surv,  1)

annot_txt <- sprintf(
  "N = %s  |  Events = %s (%.1f%%)\\nMedian event-free: %s days\\nEvent-free at 180d: %.1f%%  |  365d: %.1f%%",
  format(n_total, big.mark = ","),
  format(n_events, big.mark = ","),
  100 * n_events / n_total,
  ifelse(is.na(med_surv) | is.infinite(med_surv), ">365", round(med_surv, 0)),
  surv_180, surv_365)

ggplot(km_df %>% filter(time < 364), aes(x = time, y = surv)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#4575b4", alpha = 0.15) +
  geom_step(colour = "#4575b4", linewidth = 0.7) +
  geom_point(data = km_df %>% filter(n.cens > 0),
             aes(x = time, y = surv), shape = 3, size = 1.2,
             colour = "#4575b4", alpha = 0.6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(.5, 1), expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(0, 400, by = 60), expand = c(0.01, 0)) +
  labs(x = "Days from first PGHD upload",
       y = "Proportion event-free",
       title = NULL) +
  theme_science(base_size = 9) +
  theme(panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3)) -> p_km
ggsave(file.path(out_figures, "Figure_2_KM_time_to_event.png"),
       plot = p_km, width = 9, height = 4.5, dpi = 300, bg = "white")
p_km
```

---

## Figure 3. Step Trajectories Before First Event

```{r fig-trajectory, echo=FALSE, fig.height=6, fig.cap="Figure 3. Mean 7-day rolling average daily step count (+/- 1 SE) and mean missing days per week (+/- 1 SE) in the 5 weeks before first composite event (ED Visit, Inpatient admission, Death) or in a randomly selected reference 5-week window for the No Event group. Dashed vertical line marks the reference week (week −1 = most recent week; week −5 = earliest). Bands represent +/- 1 SE across patients. n = number of patients contributing to each group at week −1."}
# Event groups: use actual weeks before first event
event_wks <- as.data.frame(dt_analysis) %>%
  filter(first_event_type %in% c("Death","Inpatient","ED Visit"),
         !is.na(days_to_event)) %>%
  mutate(wk_to_event = -as.integer(ceiling(days_to_event / 7))) %>%
  filter(between(wk_to_event, -5L, -1L)) %>%
  select(PatientICN, first_event_type, wk_to_event,
         k_steps_7d_avg, k_steps_7d_miss_days)

# No-event group: random 5-week window
set.seed(42L)
noevent_wks <- as.data.frame(dt_analysis) %>%
  filter(first_event_type == "No Event") %>%
  group_by(PatientICN) %>%
  filter(n() >= 5L) %>%
  group_modify(~{
    n_int <- nrow(.x); si <- sample.int(max(n_int - 4L, 1L), 1L)
    .x[si:(si + 4L), ] %>% mutate(wk_to_event = seq(-5L, -1L))
  }) %>% ungroup() %>%
  select(PatientICN, first_event_type, wk_to_event,
         k_steps_7d_avg, k_steps_7d_miss_days)

traj_df <- bind_rows(event_wks, noevent_wks) %>%
  group_by(first_event_type, wk_to_event) %>%
  summarise(
    mean_steps   = mean(k_steps_7d_avg,       na.rm = TRUE) * 1000,
    se_steps     = sd(k_steps_7d_avg,         na.rm = TRUE) / sqrt(n()) * 1000,
    mean_miss    = mean(k_steps_7d_miss_days, na.rm = TRUE),
    se_miss      = sd(k_steps_7d_miss_days,   na.rm = TRUE) / sqrt(n()),
    n_pts        = n_distinct(PatientICN),
    .groups = "drop")

week_labels <- setNames(paste0("Wk ", seq(-5,-1)), seq(-5,-1))

p_steps <- traj_df %>%
  ggplot(aes(x = wk_to_event, y = mean_steps,
             color = first_event_type, group = first_event_type)) +
  geom_ribbon(aes(ymin = mean_steps - se_steps, ymax = mean_steps + se_steps,
                  fill = first_event_type), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.85) + geom_point(size = 1.8) +
  geom_vline(xintercept = -0.5, linetype = "dashed", linewidth = 0.4, color = "grey40") +
  scale_color_manual(values = event_colors, name = "Group") +
  scale_fill_manual(values  = event_colors, name = "Group") +
  scale_x_continuous(breaks = seq(-5,-1), labels = paste0("Wk ", seq(-5,-1))) +
  geom_text(data = traj_df %>% filter(wk_to_event == -1L),
            aes(label = paste0("n=", n_pts), y = mean_steps + se_steps + 150),
            size = 2.2, show.legend = FALSE) +
  labs(x = NULL, y = "Daily steps") +
  theme_science() +
  theme(axis.text.x    = element_text(angle = 35, hjust = 1),
        legend.position = "right",
        legend.text    = element_text(size = 8),
        legend.title   = element_text(size = 8))

p_miss <- traj_df %>%
  ggplot(aes(x = wk_to_event, y = mean_miss,
             color = first_event_type, group = first_event_type)) +
  geom_ribbon(aes(ymin = mean_miss - se_miss, ymax = mean_miss + se_miss,
                  fill = first_event_type), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.85) + geom_point(size = 1.8) +
  geom_vline(xintercept = -0.5, linetype = "dashed", linewidth = 0.4, color = "grey40") +
  scale_color_manual(values = event_colors, name = "Group") +
  scale_fill_manual(values  = event_colors, name = "Group") +
  scale_x_continuous(breaks = seq(-5,-1), labels = paste0("Wk ", seq(-5,-1))) +
  labs(x = "Weeks relative to event", y = "Missing days / week") +
  theme_science() +
  theme(axis.text.x     = element_text(angle = 35, hjust = 1),
        legend.position = "none")

p_traj <- p_steps / p_miss
ggsave(file.path(out_figures, "Figure_3_step_trajectories_pre_event.png"),
       plot = p_traj, width = 9, height = 6, dpi = 300, bg = "white")
p_traj
```

---

## Table 2. Predictive Model Performance

```{r table2, echo=FALSE}
col_order <- c("feature_label","algo",
               "auc_14d","pr_auc_14d","bss_14d","cal_slope_post_platt_14d","events_test_14d",
               "auc_30d","pr_auc_30d","bss_30d","cal_slope_post_platt_30d",
               "n_test_30d","events_test_30d")
present_cols <- intersect(col_order, names(m1_perf_wide))

tbl2 <- m1_perf_wide %>% select(all_of(present_cols))

# Bold max AUC in each horizon column
auc14_max <- if ("auc_14d" %in% names(tbl2)) max(tbl2$auc_14d, na.rm = TRUE) else NA
auc30_max <- if ("auc_30d" %in% names(tbl2)) max(tbl2$auc_30d, na.rm = TRUE) else NA

# Null-model (event rate) reference row for PR-AUC context
null_row <- tibble(
  feature_label            = "Null model (event rate)",
  algo                     = "",
  auc_14d                  = 0.5,
  pr_auc_14d               = round(mean(tbl2$events_test_14d / tbl2$n_test_30d, na.rm=TRUE), 4),
  bss_14d                  = 0,
  cal_slope_post_platt_14d = NA_real_,
  events_test_14d          = as.integer(round(mean(tbl2$events_test_14d, na.rm=TRUE))),
  auc_30d                  = 0.5,
  pr_auc_30d               = round(mean(tbl2$events_test_30d / tbl2$n_test_30d, na.rm=TRUE), 4),
  bss_30d                  = 0,
  cal_slope_post_platt_30d = NA_real_,
  n_test_30d               = as.integer(round(mean(tbl2$n_test_30d, na.rm=TRUE))),
  events_test_30d          = as.integer(round(mean(tbl2$events_test_30d, na.rm=TRUE)))
) %>% select(all_of(present_cols))

tbl2_fmt <- bind_rows(null_row, tbl2) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

# Pack-row indices (offset by 1 for the null row prepended)
logistic_rows <- which(tbl2_fmt$algo == "Logistic")
xgboost_rows  <- which(tbl2_fmt$algo == "XGBoost")

col_names_display <- c(
  "Feature set", "Algorithm",
  "AUC", "PR-AUC", "BSS", "Cal. slope", "Events",
  "AUC", "PR-AUC", "BSS", "Cal. slope",
  "Test N", "Events"
)[seq_along(present_cols)]

display_cols <- setdiff(present_cols, "algo")
int_cols     <- grep("^(n_test|events_test|feature_label)", display_cols, value = TRUE)
digits_vec   <- ifelse(display_cols %in% int_cols, 0L, 4L)

# Pre-format for HTML display (allows footnote superscript in one cell) 
tbl2_disp <- tbl2_fmt %>% select(-algo)
num_cols  <- display_cols[sapply(tbl2_disp[display_cols], is.numeric)]
for (.col in num_cols) {
  .decimals <- if (.col %in% int_cols) 0L else 4L
  .vals     <- tbl2_disp[[.col]]
  .fmt      <- if (.decimals == 0L) {
    formatC(as.integer(.vals), format = "d", big.mark = ",")
  } else {
    formatC(as.numeric(.vals), digits = .decimals, format = "f")
  }
  tbl2_disp[[.col]] <- ifelse(is.na(.vals), "--", .fmt)
}

col_labels_present <- setNames(
  as.list(col_names_display[-2]),
  setdiff(present_cols, "algo")
)

kbl2 <- tbl2_fmt %>%
  mutate(
    .grp = dplyr::case_when(
      algo == "Logistic" ~ "Logistic regression",
      algo == "XGBoost"  ~ "XGBoost",
      .default           = "Null model"
    )
  ) %>%
  select(-algo) %>%
  gt(groupname_col = ".grp") %>%
  cols_label(.list = col_labels_present) %>%
  tab_spanner(
    label   = "14-day horizon",
    columns = any_of(c("auc_14d","pr_auc_14d","bss_14d","cal_slope_post_platt_14d","events_test_14d"))
  ) %>%
  tab_spanner(
    label   = "30-day horizon",
    columns = any_of(c("auc_30d","pr_auc_30d","bss_30d","cal_slope_post_platt_30d","n_test_30d","events_test_30d"))
  ) %>%
  fmt_integer(columns = any_of(c("events_test_14d","n_test_30d","events_test_30d"))) %>%
  fmt_number(
    columns  = any_of(c("auc_14d","pr_auc_14d","bss_14d","cal_slope_post_platt_14d",
                         "auc_30d","pr_auc_30d","bss_30d","cal_slope_post_platt_30d")),
    decimals = 4
  ) %>%
  sub_missing(missing_text = "--") %>%
  tab_style(
    style     = cell_text(style = "italic"),
    locations = cells_row_groups(groups = "Null model")
  ) %>%
  cols_width(feature_label ~ px(192)) %>%
  tab_header(title = "Table 2. Model performance on the 30% patient-level hold-out test set.") %>%
  tab_source_note(source_note = html(paste0(
    "<b>Note.</b> ",
    "AUC = area under ROC curve; PR-AUC = precision-recall AUC; ",
    "BSS = Brier Skill Score; ",
    "BSS = 0 indicates no improvement over the null model (event rate), BSS = 1 is perfect; ",
    "Cal. slope = post-calibration slope on test set (target: 1.0); ",
    "Platt scaling fit on held-out 20% of training patients (patient-level), applied to test set for all metrics."
  ))) %>%
  tab_options(table.font.size = px(11), table.width = "auto")
kbl2
```

```{r table2-download, echo=FALSE}
if (requireNamespace("downloadthis", quietly = TRUE)) {
  tbl2_disp %>%
    setNames(col_names_display[-2]) %>%
    downloadthis::download_this(
      output_name      = "Table2_model_performance",
      output_extension = ".xlsx",
      button_label     = "Download Table 2 data (xlsx)",
      button_type      = "default",
      class            = "btn-sm"
    )
}
```


### Sensitivity Analyses

```{r sensitivity-results, echo=FALSE, results='asis'}
# Sensitivity B: stratified test-set performance by weekly data quality (30-day horizon)
sa_30d <- sa_summaries %>%
  filter(horizon == 30L) %>%
  mutate(
    feature_label = recode(feature_mode, !!!feat_labels),
    algo          = dplyr::recode(model_type, xgboost = "XGBoost", logistic = "Logistic")
  ) %>%
  select(feature_label, algo, stratum, n_weeks, events, auc, pr_auc, bss) %>%
  arrange(algo, feature_label, stratum)

sa_logistic_rows <- which(sa_30d$algo == "Logistic")
sa_xgboost_rows  <- which(sa_30d$algo == "XGBoost")

kbl_sa <- sa_30d %>%
  mutate(
    .grp = dplyr::case_when(
      algo == "Logistic" ~ "Logistic regression",
      algo == "XGBoost"  ~ "XGBoost",
      .default           = NA_character_
    )
  ) %>%
  select(.grp, feature_label, stratum, n_weeks, events, auc, pr_auc, bss) %>%
  gt(groupname_col = ".grp") %>%
  cols_label(
    feature_label = "Feature set",
    stratum       = "Stratum",
    n_weeks       = "Person-weeks",
    events        = "Events",
    auc           = "AUC",
    pr_auc        = "PR-AUC",
    bss           = "BSS"
  ) %>%
  fmt_integer(columns = c(n_weeks, events)) %>%
  fmt_number(columns  = c(auc, pr_auc, bss), decimals = 4) %>%
  cols_width(feature_label ~ px(192)) %>%
  tab_header(title = html(paste0(
    "Table S1. Sensitivity Analysis B: Test-set performance stratified by weekly data quality at the 30-day horizon. ",
    "\u2018Clean\u2019: 0\u20133 missing days per week (\u22654 days with step data). ",
    "\u2018Sparse\u2019: 4\u20136 missing days per week (1\u20133 days with step data). ",
    "No model refit - stratified evaluation within the primary model. ",
    "Comparable performance across strata indicates the model captures pre-event device-dropout signal present in sparse weeks."
  ))) %>%
  tab_options(table.font.size = px(11), table.width = "auto")
kbl_sa
```

```{r sensitivity-download, echo=FALSE}
if (requireNamespace("downloadthis", quietly = TRUE)) {
  sa_30d %>%
    downloadthis::download_this(
      output_name      = "Table_S1_sensitivity_analysis",
      output_extension = ".xlsx",
      button_label     = "Download Sensitivity Analysis data (xlsx)",
      button_type      = "default",
      class            = "btn-sm"
    )
}
```


## Table 3. Model Calibration: 14-Day and 30-Day Horizons

```{r table3-calibration, echo=FALSE}
cal_tbl_data <- m1_cal_data %>%
  mutate(
    feature_label = recode(feature_mode, !!!feat_labels),
    algo          = str_to_title(model_type),
    horizon_label = paste0(horizon, "-day")
  )

if (nrow(cal_tbl_data) > 0) {
  cal_tbl_data %>%
    mutate(
      mean_predicted = round(mean_predicted, 3),
      observed_rate  = round(observed_rate,  3),
      n              = format(n,        big.mark = ","),
      n_events       = format(n_events, big.mark = ",")
    ) %>%
    arrange(horizon_label, feature_label, algo, prob_bin) %>%
    mutate(.grp = paste0(horizon_label, " - ", feature_label, " - ", algo)) %>%
    select(
      .grp,
      `Predicted prob. bin` = prob_bin,
      N                     = n,
      `Events (n)`          = n_events,
      `Mean predicted`      = mean_predicted,
      `Observed rate`       = observed_rate
    ) %>%
    gt(groupname_col = ".grp") %>%
    tab_header(title = "Table 3. Calibration table for 14-day and 30-day horizon models. Predicted probabilities are grouped into decile-based bins; observed event rates are compared to mean predicted probabilities within each bin. Well-calibrated models show observed rates close to mean predicted values.") %>%
    cols_align(align = "left",  columns = `Predicted prob. bin`) %>%
    cols_align(align = "right", columns = c(N, `Events (n)`, `Mean predicted`, `Observed rate`)) %>%
    tab_options(
      table.font.size       = px(11),
      table.width           = "auto",
      row_group.font.weight = "bold"
    )
} else {
  cat("*Calibration data not available - ensure models have been fit.*")
}
```

```{r table3-download, echo=FALSE}
if (requireNamespace("downloadthis", quietly = TRUE) && nrow(cal_tbl_data) > 0) {
  cal_tbl_data %>%
    mutate(
      feature_label = recode(feature_mode, !!!feat_labels),
      algo          = str_to_title(model_type),
      horizon_label = paste0(horizon, "-day")
    ) %>%
    select(
      horizon_label, feature_label, algo,
      prob_bin, n, n_events, mean_predicted, observed_rate
    ) %>%
    downloadthis::download_this(
      output_name      = "Table3_calibration",
      output_extension = ".xlsx",
      button_label     = "Download Table 3 data (xlsx)",
      button_type      = "default",
      class            = "btn-sm"
    )
}
```

---

## Figure 4. ROC Curves at 30-Day Horizon

```{r fig3-roc, echo=FALSE, fig.height=4.5, fig.cap="Figure 4. Receiver operating characteristic (ROC) curves for the 30-day composite outcome prediction. Left panel: logistic regression; right panel: XGBoost. Each panel shows three curves corresponding to the three feature sets. AUC values are annotated per curve. Dotted diagonal represents chance performance."}
roc_df <- m1_roc_data_30d %>%
  mutate(
    feature_label = recode(feature_mode, !!!feat_labels),
    algo          = dplyr::recode(model_type, xgboost = "XGBoost", logistic = "Logistic"),
    auc_label     = sprintf("AUC = %.3f", auc_val),
    sens          = tpr
  )

# AUC annotation positions (pick middle of each curve)
auc_annot <- roc_df %>%
  dplyr::group_by(feature_label, algo) %>%
  dplyr::slice(round(dplyr::n() / 2)) %>%
  dplyr::ungroup()

make_roc_panel <- function(algo_name) {
  d <- roc_df %>% filter(algo == algo_name)
  a <- auc_annot %>% filter(algo == algo_name)
  ggplot(d, aes(x = fpr, y = sens, color = feature_label, group = feature_label)) +
    geom_line(linewidth = 0.9) +
    geom_abline(slope = 1, intercept = 0, linetype = "dotted",
                linewidth = 0.3, color = "grey60") +
    ggrepel::geom_label_repel(
      data = a %>% group_by(feature_label) %>% slice(1),
      aes(label = auc_label), size = 2.4, box.padding = 0.3,
      segment.size = 0.3, show.legend = FALSE,
      family = "Helvetica") +
    scale_color_manual(values = feat_colors, name = "Feature set") +
    scale_x_continuous(labels = percent_format(accuracy = 1),
                       limits = c(0, 1), expand = expansion(mult = .02)) +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       limits = c(0, 1), expand = expansion(mult = .02)) +
    labs(title = algo_name, x = "1 – Specificity", y = "Sensitivity") +
    theme_science() +
    theme(legend.position = "bottom",
          legend.text  = element_text(size = 7.5),
          legend.title = element_text(size = 7.5),
          plot.title   = element_text(size = 9, face = "bold"))
}

# ── AUC summary dot-and-whisker panel ────────────────────────────────────────────
if (exists("m1_summaries") && nrow(m1_summaries) > 0L) {
  p_auc_summary <- m1_summaries %>%
    mutate(
      feature_label = recode(feature_mode, !!!feat_labels),
      algo          = dplyr::recode(model_type, xgboost = "XGBoost", logistic = "Logistic"),
      horizon_label = paste0(horizon, "-day")
    ) %>%
    ggplot(aes(x = feature_label, y = auc,
               color = horizon_label, shape = algo,
               group = interaction(algo, horizon_label))) +
    geom_point(size = 3, position = position_dodge(0.35)) +
    geom_line(aes(group = interaction(algo, horizon_label)),
              position = position_dodge(0.35), linewidth = 0.6) +
    scale_color_manual(values = c("14-day" = "#4575b4", "30-day" = "#d73027"),
                       name = "Horizon") +
    scale_shape_manual(values = c("Logistic" = 16L, "XGBoost" = 17L),
                       name = "Algorithm") +
    scale_y_continuous(limits = c(0.5, 1),
                       labels = scales::number_format(accuracy = 0.01)) +
    labs(title = "AUC by Feature Set & Horizon",
         x = "Feature set", y = "AUC (ROC)") +
    theme_science() +
    theme(axis.text.x    = element_text(angle = 20, hjust = 1),
          legend.position = "bottom",
          legend.text    = element_text(size = 7.5),
          legend.title   = element_text(size = 7.5))
} else {
  p_auc_summary <- ggplot() + theme_void() + labs(title = "AUC summary unavailable")
}

# Install ggrepel if needed
if (!requireNamespace("ggrepel", quietly = TRUE))
  try(install.packages("ggrepel", repos = "https://cloud.r-project.org", quiet = TRUE), silent = TRUE)
if (!requireNamespace("ggrepel", quietly = TRUE)) library(ggrepel)

p_roc <- tryCatch({
  library(ggrepel)
  make_roc_panel("Logistic") + make_roc_panel("XGBoost") + p_auc_summary +
    plot_layout(ncol = 3, guides = "collect") & theme(legend.position = "bottom")
}, error = function(e) {
  # Fallback without ggrepel labels
  roc_df %>%
    ggplot(aes(x = fpr, y = sens, color = feature_label)) +
    geom_line(linewidth = 0.9) +
    geom_abline(slope=1,intercept=0,linetype="dotted",linewidth=0.3, color="grey60") +
    facet_wrap(~algo) +
    scale_color_manual(values = feat_colors, name = "Feature set") +
    scale_x_continuous(labels = percent_format(accuracy=1)) +
    scale_y_continuous(labels = percent_format(accuracy=1)) +
    labs(x = "1 – Specificity", y = "Sensitivity") +
    theme_science() +
    theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
})
ggsave(file.path(out_figures, "Figure_4_ROC_curves_30d.png"),
       plot = p_roc, width = 9, height = 4.5, dpi = 300, bg = "white")
p_roc
```

---

## Figure 5. XGBoost Feature Importance: Steps + Static Model

```{r fig4-importance, echo=FALSE, fig.height=8, fig.cap="Figure 5. Top 20 XGBoost feature importances (Gain): Steps + Static, 14-Day and 30-Day Horizons. Bars filled by feature category (colorblind-friendly Set2 palette). Gain = fractional contribution to total information gain across all trees. Labels show Gain x 100."}
# Shared label helper - used by this chunk and the logistic regression chunk
feat_display_names <- c(
  # PGHD / step features
  baseline_k_steps_avg         = "Baseline avg. daily steps",
  baseline_k_steps_miss_days   = "Baseline avg. missing days/week",
  k_steps_7d_avg               = "7-day avg. daily steps",
  k_steps_7d_sd                = "7-day SD of daily steps",
  k_steps_7d_miss_days         = "Missing days/week (7-day)",
  k_steps_7d_all_missing       = "All days missing flag (7-day)",
  k_steps_7d_pct_from_baseline = "% change from baseline steps",
  k_steps_7d_avg_pctdiff       = "% change vs. prior 3-week avg.",
  low_activity_days_7d         = "Low-activity days/week (7-day)",
  low_activity_streak          = "Consecutive low-activity days",
  k_steps_14d_avg              = "14-day avg. daily steps",
  k_steps_14d_sd               = "14-day SD of daily steps",
  k_steps_21d_avg              = "21-day avg. daily steps",
  k_steps_21d_sd               = "21-day SD of daily steps",
  # Vitals & labs
  bmi_fitted_mean              = "BMI, fitted mean (kg/m\u00b2)",
  bmi_slope                    = "BMI trend slope",
  bmi_fitted_last              = "BMI, most recent fitted (kg/m\u00b2)",
  map_fitted_mean              = "MAP, fitted mean (mmHg)",
  map_fitted_last              = "MAP, most recent fitted (mmHg)",
  map_slope                    = "MAP trend slope",
  recent_pulse                 = "Heart rate, most recent (bpm)",
  bun_value_mean               = "BUN, mean (mg/dL)",
  bun_value_last               = "BUN, most recent (mg/dL)",
  albumin_value_mean           = "Albumin, mean (g/dL)",
  albumin_value_last           = "Albumin, most recent (g/dL)",
  leuk_value_mean              = "WBC, mean (10\u00b3/\u00b5L)",
  leuk_value_last              = "WBC, most recent (10\u00b3/\u00b5L)",
  # Demographics
  Age                          = "Age (years)",
  # Gender (ref = Male)
  gender_Female                = "Female (Male)",
  gender_Unknown               = "Unknown Sex (Male)",
  # Race (ref = White)
  race_BlackOrAfricanAmerican  = "Black or African American (White)",
  race_AmericanIndian          = "American Indian or Alaska Native (White)",
  race_Asian                   = "Asian (White)",
  race_PacificIslander         = "Native Hawaiian or Pacific Islander (White)",
  race_Multiracial             = "Multiracial (White)",
  race_Unknown                 = "Unknown Race (White)",
  # Ethnicity (ref = Not Hispanic or Latino)
  eth_HispanicOrLatino         = "Hispanic or Latino (Not Hispanic or Latino)",
  eth_Unknown                  = "Unknown Ethnicity (Not Hispanic or Latino)",
  # Marital status (ref = Married)
  marital_Divorced             = "Divorced (Married)",
  marital_NeverMarried         = "Never Married (Married)",
  marital_Widowed              = "Widowed (Married)",
  marital_Separated            = "Separated (Married)",
  marital_Unknown              = "Unknown Marital Status (Married)",
  # Rurality (ref = Urban)
  gisurh_Rural                 = "Rural (Urban)",
  gisurh_HighlyRural           = "Highly Rural (Urban)",
  gisurh_Insular               = "Insular (Urban)",
  gisurh_Unknown               = "Unknown Rurality (Urban)",
  # VA Priority Group - individual groups (ref = Group 1)
  pg_G2                        = "Priority Group 2 (Group 1)",
  pg_G3                        = "Priority Group 3 (Group 1)",
  pg_G4                        = "Priority Group 4 (Group 1)",
  pg_G5                        = "Priority Group 5 (Group 1)",
  pg_G6                        = "Priority Group 6 (Group 1)",
  pg_G7                        = "Priority Group 7 (Group 1)",
  pg_G8                        = "Priority Group 8 (Group 1)",
  pg_Unknown                   = "Unknown Priority Group (Group 1)",
  # Season (ref = Fall)
  season_Spring                = "Spring (Fall)",
  season_Summer                = "Summer (Fall)",
  season_Winter                = "Winter (Fall)",
  days_obs_remaining           = "Days remaining in follow-up",
  # Healthcare utilization
  py_ed                        = "ED visit, prior year",
  py_urgent                    = "Urgent care visit, prior year",
  py_appt_no_show              = "Appointment no-show, prior year",
  py2_inpat_any                = "Any inpatient admission, 2-year",
  py2_inpat_med_surg           = "Med./surgical admission, 2-year",
  py2_inpat_mental_health      = "Mental health admission, 2-year",
  py2_inpat_other              = "Other inpatient admission, 2-year",
  py2_inpat_nursing_home       = "Nursing home admission, 2-year",
  py2_IVC_n                    = "Community care stays, count (2-year)",
  py2_IVC_los                  = "Community care LOS, days (2-year)",
  py2_IVC                      = "Community care stay, any (2-year)",
  py2_mwi_score                = "Multimorbidity Weighted Index (MWI)",
  # Comorbidity flags
  py2_congestive_heart_failure = "Congestive heart failure, 2-year",
  py2_heart_arrhythmias        = "Cardiac arrhythmia, 2-year",
  py2_drug_alcohol_dependence  = "Substance use disorder, 2-year",
  py2_metastatic_cancer        = "Metastatic cancer, 2-year",
  py2_lung_severe_cancers      = "Lung/severe cancer, 2-year",
  py2_dialysis_status          = "Dialysis, 2-year",
  py2_housing_instability      = "Housing instability, 2-year",
  py2_financial_sdoh           = "Financial hardship, 2-year",
  py2_acute_renal_failure      = "Acute renal failure, 2-year",
  py2_artificial_openings      = "Artificial openings, 2-year",
  py2_chronic_pancreatitis     = "Chronic pancreatitis, 2-year",
  py2_chronic_ulcer_skin       = "Chronic skin/ulcer condition, 2-year",
  py2_drug_alcohol_psychosis   = "Drug/alcohol-induced psychosis, 2-year",
  py2_pregnancy                = "Pregnancy/peripartum condition, 2-year",
  py2_nonspecific_psychosocial = "Nonspecific psychosocial condition, 2-year",
  py2_dialysis_access          = "Dialysis access, 2-year",
  cci_score                    = "Charlson Comorbidity Index (CCI)",
  # Social history
  mst_Yes                      = "Military Sexual Trauma: Yes (No)",
  current_smoker               = "Current smoker",
  never_smoker                 = "Never smoker"
)

  apply_feat_labels <- function(term_vec) {
    # Cleans recipe-encoded level suffixes: replaces dots with spaces (e.g.
    # "Black.or.African.American" -> "Black or African American") and converts
    # all-uppercase strings to title case (e.g. "MALE" -> "Male").
    clean_lv <- function(x) {
      x <- gsub(".", " ", x, fixed = TRUE)
      dplyr::if_else(x == toupper(x), tools::toTitleCase(tolower(x)), x)
    }
    # All labels are pre-defined in feat_display_names; fall back to raw column name
    case_when(
      term_vec %in% names(feat_display_names) ~ feat_display_names[term_vec],
      TRUE ~ term_vec
    )
  }

  # Load pre-computed XGBoost importance (saved in target block)
  imp_combined <- m1_xgb_importance %>%
    filter(feature_mode == "model3_steps_and_static") %>%
    group_by(horizon) %>%
    arrange(desc(gain)) %>%
    slice_head(n = 20) %>%
    ungroup() %>%
    mutate(
      horizon_label = if_else(horizon == 14L, "14-Day Horizon", "30-Day Horizon"),
      Category      = classify_feature(term),
      Gain          = round(gain, 5),
      Feature_label = apply_feat_labels(term),
      horizon_label = factor(horizon_label, levels = c("14-Day Horizon", "30-Day Horizon"))
    )

  # ── Save full XGBoost importance table (both horizons) to xlsx ────────────────
  xgb_all_imp <- m1_xgb_importance %>%
    mutate(Horizon = if_else(horizon == 14L, "14-Day", "30-Day")) %>%
    arrange(Horizon, desc(gain)) %>%
    mutate(
      Category      = classify_feature(term),
      Feature_label = apply_feat_labels(term)
    ) %>%
    select(Horizon, Feature_label, Feature_raw = term, Category,
           Gain = gain, Cover = cover, Frequency = frequency)
  writexl::write_xlsx(xgb_all_imp, file.path(out_data, "xgboost_feature_importance_14d_30d.xlsx"))

  # Per-facet y-axis ordering via tidytext
  if (!requireNamespace("tidytext", quietly = TRUE))
    try(install.packages("tidytext", repos = "https://cloud.r-project.org", quiet = TRUE), silent = TRUE)
  if (requireNamespace("tidytext", quietly = TRUE)) {
    imp_combined <- imp_combined %>%
      mutate(Feature_ord = tidytext::reorder_within(Feature_label, Gain, horizon_label))
    y_scale_xgb <- tidytext::scale_y_reordered()
  } else {
    imp_combined <- imp_combined %>%
      mutate(Feature_ord = reorder(Feature_label, Gain))
    y_scale_xgb <- scale_y_discrete()
  }

  if (nrow(imp_combined) == 0) {
    cat("No XGBoost results available for Steps + Static.")
  } else {
    ggplot(imp_combined, aes(x = Gain, y = Feature_ord, fill = Category)) +
      geom_col(width = 0.7) +
      geom_text(aes(label = round(Gain * 100, 2)), hjust = -0.1, size = 2.5) +
      scale_fill_brewer(palette = "Set2", name = "Feature category") +
      scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
      y_scale_xgb +
      facet_wrap(~horizon_label, ncol = 2, scales = "free_y") +
      labs(
        title = "",
        x     = "Gain (fraction of total information gain)",
        y     = NULL
      ) +
      theme_science() +
      theme(
        axis.text.y     = element_text(size = 8),
        legend.position = "bottom",
        legend.text     = element_text(size = 8),
        legend.title    = element_text(size = 8),
        plot.title      = element_text(size = 9),
        strip.text      = element_text(size = 9, face = "bold")
      ) -> p_xgb_imp
    ggsave(file.path(out_figures, "Figure_5_XGBoost_feature_importance.png"),
           plot = p_xgb_imp, width = 9, height = 8, dpi = 300, bg = "white")
    p_xgb_imp
  }
```

---

## Figure 6. Logistic Regression Coefficients: Steps + Static Model

```{r fig5-logistic-coefs, echo=FALSE, fig.height=8, fig.cap="Figure 6. Top 20 elastic net–regularized logistic regression coefficients (OR scale, per 1 SD for continuous predictors): Steps + Static, 14-Day and 30-Day Horizons. Colored by feature category (Set2 palette). These are penalized estimates and should be interpreted as directional associations reflecting relative feature importance within the model, not formal hypothesis tests."}

# Load pre-computed logistic coefficients
m1_log_coefs <- arrow::read_parquet("data/output/m1_log_coefs.parquet")

coef_input <- m1_log_coefs %>%
  filter(feature_mode == "model3_steps_and_static",
         term != "(Intercept)", !is.na(odds_ratio))

if (nrow(coef_input) > 0) {

  coef_combined <- coef_input %>%
    group_by(horizon) %>%
    arrange(desc(abs(statistic))) %>%
    slice_head(n = 20) %>%
    ungroup() %>%
    mutate(
      horizon_label = if_else(horizon == 14L, "14-Day Horizon", "30-Day Horizon"),
      Category      = classify_feature(term),
      Feature_label = apply_feat_labels(term),
      or_label      = sprintf("%.3f", odds_ratio),
      horizon_label = factor(horizon_label, levels = c("14-Day Horizon", "30-Day Horizon"))
    )

  # ── Save full logistic coefficient table (both horizons) to xlsx ──────────────
  log_all_coef <- m1_log_coefs %>%
    filter(term != "(Intercept)") %>%
    mutate(
      Horizon       = if_else(horizon == 14L, "14-Day", "30-Day"),
      Category      = classify_feature(term),
      Feature_label = apply_feat_labels(term)
    ) %>%
    arrange(Horizon, desc(abs(estimate))) %>%
    select(Horizon, Feature_label, Feature_raw = term, Category,
           Log_Odds = estimate, SE = std.error, Z = statistic,
           P_value = p.value, Odds_Ratio = odds_ratio)
  writexl::write_xlsx(log_all_coef, file.path(out_data, "logistic_coefficients_14d_30d.xlsx"))

  # Per-facet y-axis ordering
  if (requireNamespace("tidytext", quietly = TRUE)) {
    coef_combined <- coef_combined %>%
      mutate(Feature_ord = tidytext::reorder_within(Feature_label, abs(estimate), horizon_label))
    y_scale_log <- tidytext::scale_y_reordered()
  } else {
    coef_combined <- coef_combined %>%
      mutate(Feature_ord = reorder(Feature_label, abs(estimate)))
    y_scale_log <- scale_y_discrete()
  }

  if (nrow(coef_combined) == 0) {
    cat("No logistic results available for Steps + Static.")
  } else {
    ggplot(coef_combined,
           aes(x = odds_ratio, y = Feature_ord, color = Category)) +
      geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40", linewidth = 0.5) +
      geom_point(size = 2) +
      geom_text(aes(label = or_label), hjust = -0.25, size = 2.5, color = "black") +
      scale_color_brewer(palette = "Set2", name = "Feature category") +
      scale_x_continuous(limits = c(0.75, 1.75), expand = expansion(mult = c(0.02, 0.08))) +
      y_scale_log +
      facet_wrap(~horizon_label, ncol = 2, scales = "free_y") +
      labs(
        title = "",
        x     = "Regularized coefficient (OR scale, per 1 SD for continuous predictors)",
        y     = NULL
      ) +
      theme_science() +
      theme(
        axis.text.y     = element_text(size = 8),
        legend.position = "bottom",
        legend.text     = element_text(size = 8),
        legend.title    = element_text(size = 8),
        plot.title      = element_text(size = 9),
        strip.text      = element_text(size = 9, face = "bold")
      ) -> p_log_coef
    ggsave(file.path(out_figures, "Figure_6_Logistic_OR_coefficients.png"),
           plot = p_log_coef, width = 9, height = 8, dpi = 300, bg = "white")
    p_log_coef
  }
} # end if (nrow(coef_input) > 0)
```

*Note: Coefficients are elastic net–penalized estimates (glmnet, Steps + Static model). Penalization shrinks estimates toward zero and the plot carries no inferential standard errors or significance thresholds; values reflect the direction and relative magnitude of each feature's contribution within the regularized model. A full coefficient table is saved to `logistic_coefficients_14d_30d.xlsx`.*

---

*Analysis completed: `r format(Sys.time(), "%Y-%m-%d %H:%M")`*  
*Seed: 2026 | R version: `r R.version$version.string`*
